```{r load_libs, echo=FALSE, message=FALSE, warning=FALSE}
# clean environment
rm(list = ls())  
# load packages
library(sf)
library(tidyverse)
library(skimr)
library(tidymodels)
```

```{r load_data, echo=FALSE, message=FALSE, warning=FALSE}
# Shetland BBS
survey_BBS_2002_2018     <- st_read("/Users/anthony/Documents/GitHub/Waders/data/Shetland_BBS/BBSgrid2002_2018.dbf", quiet=T)

```


# Results

## Explororatory Data Analysis of SBBS survey data

Where relevant a protocol for exploratory data analysis was followed [@Zuur2010-kp] to ensure that before any problems in the structure of the data are identified prior to undertaking any statistical analysis.

### Outliers

The *Cleveland dot plot* [@Zuur2010-kp] is a chart in which the row number of an observation is plotted versus the observation variable, thereby providing a more detailed view of individual observations than a boxplot. Points that stick out on the right-hand side, or on the left-hand side, are observed values that are considerably larger, or smaller, than the majority of the observations. Figure \@ref(fig:countDotPlot) appears to show that there are no major outliers across all species, but that there are many counts equal to zero indicating that the data might be zero-inflated.

```{r countDotPlot, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Cleveland dot plot of species counts in Shetland BBS data from 2002 to 2018 '}

survey_BBS_2002_2018 %>%
  # pivot long by species
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  # Order data by observation position in the data set
  ggplot(aes(x=count, y=seq(1, length(count)))) +
  # Make the chart points transparent 
  geom_point(alpha=0.3) +
  labs(x = "Species count per BBS 1km square", 
       y = "Order of data") +
  theme_bw() +
  # Plot for each species
  facet_wrap(~species, ncol = 2)
```
### Testing for normality

A large number of statistical regression techniques assume normality. Visualising the SBBS count data as a histogram can help assess if it is normally distributed. This is shown in the plot in Figure \@ref(fig:normality).

```{r normality, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Histogram of SBBS count data across all years, by species '}
survey_BBS_2002_2018 %>%
  # pivot long by species
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  ggplot(aes(x = count)) + 
  # Set the histogram binwidth to 0.25g
  geom_histogram(color = "black", binwidth = 0.25) +
  theme_minimal() +
  labs(x = "Species counts", y = "Frequency") +
  facet_wrap(~species, ncol=2)

```
Inorder to validate the outcome of the plots in Figure \@ref(fig:normality) a significance test was undertaken and the results are shown in Table \@ref(tab:normSig).

```{r normSig, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Shapiro-Wilk normality test of Shetland SBBS species counts, across all year'}
# load library for stats test
library(car)

survey_BBS_2002_2018 %>%
  # pivot long by species counts
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  # map over species
  nest(-species) %>%   
  # run fligner test across each species, count ~ year
  mutate(tidy = map(data, ~shapiro.test(.$count) %>% tidy)) %>% 
  # pull out tidyed output
  unnest(tidy) %>%
  # remove cols we dont need
  select(-data,-method) %>%
  # knit table
  knitr::kable(
    col.names = c("Species","W","p-value"),
    booktabs = TRUE
  )

```

The p-value for each species in \@ref(tab:normSig) is << 0.05. This suggests that the count data for all species are significantly different from the normal distribution.

### Poisson distribution and zero inflation

The histograms of species counts in Figure \@ref(fig:normality) suggest that count data is poisson distribution. Also there are a significant number of zeros in the count data, for all species. This suggests that the zero-inflation poisson distribution describes the data. Table \@ref(tab:zipTest) below shows the results of a significance test [@Van_den_Broek1995-ml] for zero inflation in a poisson distribution.

```{r zipTest, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Zero inflation poisson distribution test for Shetland SBBS count data, across all years and species'}

zi.test <- function(data){
  data <- data[!is.na(data)]
  lambda_est <- mean(data)
  p0_tilde <- exp(-lambda_est)
  p0_tilde
  n0 <- sum(1*(!(data >0)))
  n <- length(data)

  #now lets perform the JVDB score test
  numerator <- (n0 -n*p0_tilde)^2
  denominator <- n*p0_tilde*(1-p0_tilde) - n*lambda_est*(p0_tilde^2)

  test_stat <- numerator/denominator

  pvalue <- pchisq(test_stat,df=1, ncp=0, lower.tail=FALSE)
  pvalue

  return(list(
    # expected number of observations expected to be zero
    exp.0 = n*p0_tilde, 
    # number of zero observations
    obs.0 = n0,
    # chi squared stat
    chi.sq=test_stat,
    #p-value
    p.value=pvalue ))


}
# Run the test on each species
zip_test_res <- survey_BBS_2002_2018 %>%
  # pivot long by species counts
  pivot_longer(cols=OC:SN,names_to = "species", values_to = "count") %>%
  dplyr::select(species,count) %>%
  # create species subsets
  split(.$species) %>%
  map_dfr(
    # run zip test over species count
    ~zi.test(.x$count), 
    .id = toString(.$species)
    )

# create table of results
zip_test_res %>%
  # knit table
  knitr::kable(
    col.names = c("Species","Expected zeros","Zeros observed", "Chi squared","p-value"),
    booktabs = TRUE
  )

```

All results have a significant statistical significance (p<0.05) and therefore the count distribution across species is assumed to be a zero-inflated poisson process. The statistical modelling methods used on the data must support a poisson distribution and zero inflation where possible.

### Homogeniety of variance

Homogeneity of variance within the data is an important assumption in analysis of variance (ANOVA) and other regression-related models. The series of boxplots in Figure \@ref(fig:homoVariance) show how counts across all surveyed BBS squares vary across years 2002 to 2018, for each breeding wader species.

```{r homoVariance, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp=1.2, fig.cap='Box plot showing variance of counts across all surveyed Shetland BBS squares and all years, by species'}
survey_BBS_2002_2018 %>%
  # pivot long by species
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  ggplot(aes(x=YEAR, y=count)) +
    geom_boxplot(aes(group=YEAR)) +
    # Facet by year
    facet_wrap(~species, ncol = 1) +
    theme_bw() 

```

To test the homogeneity of variance of species counts between years, for each species, we can apply the Fligner-Killeen test. This is used as the count data are shown to be non-normal. Table \@ref(tab:fkTest) shows the results of the test applied to the Shetland BBS data. For p-values > 0.05 the data variance are homogeneous.

```{r fKTest, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center'}
survey_BBS_2002_2018 %>%
  # pivot long by species counts
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  # map over species
  nest(-species) %>%   
  # run fligner test across each species, count ~ year
  mutate(tidy = map(data, ~fligner.test(count ~ as.factor(YEAR), data = .) %>% tidy)) %>% 
  # pull out tidyed output
  unnest(tidy) %>%
  # remove cols we dont need
  select(-data,-method) %>%
  # knit table
  knitr::kable(
    col.names = c("Species","Chi-squared","p-value","df"),
    booktabs = TRUE,
    caption = "Fligner-Killeen test of homogeneity of variance for Shetland SBBS species counts, across all years"
  )
```


Lapwing, Redshank and Snipe variances are inhomogeneous according to the test results in Table \@ref(tab:fKTest). The solution to heterogeneity of
variance is to transform the response variable to stabilize the variance year-on-year, or applying statistical regression techniques that do not require homogeneity. 

### Changes over two periods




### Count data distribution

### Spatial view of survey sites


## Survey  Bootstrap

## Improved grassland classification
Spectral response charts

## Landcover histograms
By EUNIS type


## Density plots

## Environmental covariate response

## IT covariate response

## Abundance spatial distribution

## Abundance population trends