## Explororatory Data Analysis of SBBS survey data

Where relevant a protocol for exploratory data analysis was followed [@Zuur2010-kp] to ensure that before any problems in the structure of the data are identified prior to undertaking any statistical analysis.

### Survey effort over time

The spatial location of surveyed squares is shown in Figure \@ref(fig:spatSum). It seems that there has been ongoing surveying effort in the south and central mainland and on the islands of Unst, Bressay and Noss, but less coverage elsewhere.

```{r spatSum,fig.asp = 0.8, fig.width = 8,message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Number of years a SBBS 1km square was survyered (n) between 2002 and 2019'}
# Read coastal features for Shetland as spatial features from osm dataset
shetland_lines_sf <- st_read("~/Documents/GitHub/shetlandwaders/data/shetland_osm.osm", layer = "lines" , quiet = T)
shetland_polys_sf <- st_read("~/Documents/GitHub/shetlandwaders/data/shetland_osm.osm", layer = "multipolygons", quiet = T)

# Generate number of visits per BBS square
spat_summary <- survey_BBS_2002_2018 %>% 
  group_by(PLAN_NO) %>% 
  summarise(n=n())

ggplot(data = spat_summary) +
  geom_sf() +
  # Add the coastal features that are lines
  geom_sf(data = shetland_lines_sf, lwd=0.2, color = "gray50") +
  # Add the coastal features that are polygons
  geom_sf(data = shetland_polys_sf, lwd=0.2, color = "gray50", fill = NA) +
  geom_sf(aes(fill = n),lwd = 0, show.legend = TRUE) +
  scale_fill_viridis_c(trans = "sqrt") +
  coord_sf(crs=27700, xlim = c(413618, 467722), ylim = c(1109567, 1218626)) +
  theme_bw()
```

### Outliers

The *Cleveland dot plot* [@Zuur2010-kp] is a chart in which the row number of an observation is plotted versus the observation variable, thereby providing a more detailed view of individual observations than a boxplot. Points that stick out on the right-hand side, or on the left-hand side, are observed values that are considerably larger, or smaller, than the majority of the observations. Figure \@ref(fig:countDotPlot) appears to show that there are no major outliers across all species, but that there are many counts equal to zero indicating that the data might be zero-inflated.

```{r countDotPlot, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Cleveland dot plot of species counts in Shetland BBS data from 2002 to 2018 '}

survey_BBS_2002_2018 %>%
  # pivot long by species
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  # Order data by observation position in the data set
  ggplot(aes(x=count, y=seq(1, length(count)))) +
  # Make the chart points transparent 
  geom_point(alpha=0.3) +
  labs(x = "Species count per BBS 1km square", 
       y = "Order of data") +
  theme_bw() +
  # Plot for each species
  facet_wrap(~species, ncol = 2)
```
### Testing for normality

A large number of statistical regression techniques assume normality. Visualising the SBBS count data as a histogram can help assess if it is normally distributed. This is shown in the plot in Figure \@ref(fig:normality).

```{r normality, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Histogram of SBBS count data across all years, by species'}
survey_BBS_2002_2018 %>%
  # pivot long by species
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  ggplot(aes(x = count)) + 
  # Set the histogram binwidth to 0.25g
  geom_histogram(color = "black", binwidth = 0.25) +
  theme_minimal() +
  labs(x = "Species counts", y = "Frequency") +
  facet_wrap(~species, ncol=2)

```
Inorder to validate the outcome of the plots in Figure \@ref(fig:normality) a significance test was undertaken and the results are shown in Table \@ref(tab:normSig).

```{r normSig, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Shapiro-Wilk normality test of Shetland SBBS species counts, across all year'}
# load library for stats test
library(car)

survey_BBS_2002_2018 %>%
  # pivot long by species counts
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  # map over species
  nest(-species) %>%   
  # run fligner test across each species, count ~ year
  mutate(tidy = map(data, ~shapiro.test(.$count) %>% tidy)) %>% 
  # pull out tidyed output
  unnest(tidy) %>%
  # remove cols we dont need
  dplyr::select(-data,-method) %>%
  # knit table
  knitr::kable(
    col.names = c("Species","W","p-value"),
    booktabs = TRUE
  )

```

The p-value for each species in \@ref(tab:normSig) is << 0.05. This suggests that the count data for all species are significantly different from the normal distribution.

### Poisson distribution and zero inflation

The histograms of species counts in Figure \@ref(fig:normality) suggest that count data is poisson distribution. Also there are a significant number of zeros in the count data, for all species. This suggests that the zero-inflation poisson distribution describes the data. Table \@ref(tab:zipTest) below shows the results of a significance test [@Van_den_Broek1995-ml] for zero inflation in a poisson distribution.

```{r zipTest, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Zero inflation poisson distribution test for Shetland SBBS count data, across all years and species'}

zi.test <- function(data){
  data <- data[!is.na(data)]
  lambda_est <- mean(data)
  p0_tilde <- exp(-lambda_est)
  p0_tilde
  n0 <- sum(1*(!(data >0)))
  n <- length(data)

  #now lets perform the JVDB score test
  numerator <- (n0 -n*p0_tilde)^2
  denominator <- n*p0_tilde*(1-p0_tilde) - n*lambda_est*(p0_tilde^2)

  test_stat <- numerator/denominator

  pvalue <- pchisq(test_stat,df=1, ncp=0, lower.tail=FALSE)
  pvalue

  return(list(
    # expected number of observations expected to be zero
    exp.0 = n*p0_tilde, 
    # number of zero observations
    obs.0 = n0,
    # chi squared stat
    chi.sq=test_stat,
    #p-value
    p.value=pvalue ))
}
# Run the test on each species
zip_test_res <- survey_BBS_2002_2018 %>%
  # pivot long by species counts
  pivot_longer(cols=OC:SN,names_to = "species", values_to = "count") %>%
  dplyr::select(species,count) %>%
  # create species subsets
  split(.$species) %>%
  map_dfr(
    # run zip test over species count
    ~zi.test(.x$count), 
    .id = toString(.$species)
    )

# create table of results
zip_test_res %>%
  # knit table
  knitr::kable(
    col.names = c("Species","Expected zeros","Zeros observed", "Chi squared","p-value"),
    booktabs = TRUE
  )

```

All results have a significant statistical significance (p<0.05) and therefore the count distribution across species is assumed to be a zero-inflated poisson process. The statistical modelling methods used on the data must support a poisson distribution and zero inflation where possible.

### Homogeniety of variance

Homogeneity of variance within the data is an important assumption in analysis of variance (ANOVA) and other regression-related models. The series of boxplots in Figure \@ref(fig:homoVariance) show how counts across all surveyed BBS squares vary across years 2002 to 2018, for each breeding wader species.

```{r homoVariance, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp=1.2, fig.cap='Box plot showing variance of counts across all surveyed Shetland BBS squares and all years, by species'}
survey_BBS_2002_2018 %>%
  # pivot long by species
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  ggplot(aes(x=YEAR, y=count)) +
    geom_boxplot(aes(group=YEAR)) +
    # Facet by year
    facet_wrap(~species, ncol = 1) +
    theme_bw() 

```

To test the homogeneity of variance of species counts between years, for each species, we can apply the Fligner-Killeen test. This is used as the count data are shown to be non-normal. Table \@ref(tab:fKTest) shows the results of the test applied to the Shetland BBS data. For p-values > 0.05 the data variance are homogeneous.

```{r fKTest, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center'}
survey_BBS_2002_2018 %>%
  # pivot long by species counts
  pivot_longer(cols=OC:SN,names_to = "species",values_to = "count") %>%
  # map over species
  nest(-species) %>%   
  # run fligner test across each species, count ~ year
  mutate(tidy = map(data, ~fligner.test(count ~ as.factor(YEAR), data = .) %>% tidy)) %>% 
  # pull out tidyed output
  unnest(tidy) %>%
  # remove cols we dont need
  dplyr::select(-data,-method) %>%
  # knit table
  knitr::kable(
    col.names = c("Species","Chi-squared","p-value","df"),
    booktabs = TRUE,
    caption = "Fligner-Killeen test of homogeneity of variance for Shetland SBBS species counts, across all years"
  )
```


Lapwing, Redshank and Snipe variances are heterogeneous according to the test results in Table \@ref(tab:fKTest). The solution to heterogeneity of
variance is to transform the response variable to stabilize the variance year-on-year, or applying statistical regression techniques that do not require homogeneity. 

### Status of surveys between 2002 and 2019

```{r load_data_pop_chg, message=FALSE, warning=FALSE, echo=FALSE}
# load BBS data for all species and years
curlew        <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Curlew.xlsx", col_names = T)
lapwing       <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Lapwing.xlsx", col_names = T)
oystercatcher <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/oystercatcher.xlsx", col_names = T)
redshank      <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Redshank.xlsx", col_names = T)
snipe         <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Snipe.xlsx", col_names = T)

# join all BBS data into one dataframe
bbs_df <- rbind(curlew,
                lapwing,
                oystercatcher,
                redshank,
                snipe)

# Find squares where no survey undertaken in first and/or second time periods, and remove
na_2002_2010 <- bbs_df %>% filter_at(vars(`2002`:`2010`), all_vars(is.na(.)))
na_2011_2019 <- bbs_df %>% filter_at(vars(`2011`:`2019`), all_vars(is.na(.)))

# Transform data format
bbs_df <- bbs_df %>%
  # We aren't interested in non-breeding birds that are present.
  mutate_if(is.character, ~ifelse(.=="P", 0, .)) %>%
  mutate_if(is.character, ~ifelse(.=="p", 0, .)) %>%
  # Pivot data into long format
  pivot_longer(cols = c(`2002`:`2019`), names_to = "Year", values_to = "Count") %>%
  # Change type for number columns
  mutate_at(.vars = c("Year", "Count"), ~as.numeric(.))

# Load ordnance survey 1km grid squares
sbbs_sq_sfc <- st_read("~/Documents/GitHub/shetlandwaders/data/osgb_1km_grid/OSGB_Grid_1km_Shetland.shp", quiet=T)

# Create a df with species count averages, by square, over a given date range
calc_mn_pop <- function(date_range){
  pop_mn <- bbs_df %>% 
     # Select years within given date range
     filter(Year %in% date_range) %>%
     # pivot so that each species count has a separate column
     pivot_wider(names_from = Species, values_from = Count, values_fn = list(Count = mean)) %>%
     # Calculate mean count across all years, by species
     group_by(PLAN_NO) %>%
     summarise_at(vars(-Year), ~mean(., na.rm=TRUE))
  
  # Return result
  return(pop_mn)
}

# Calculate mean population for each wader 
# square, for both time spans
pop_2011_19 <- calc_mn_pop(c(2011:2019))
pop_2002_10 <- calc_mn_pop(c(2002:2010))

# Assigin categorical population change to each BBS square
pop_chg_df <- map2_dfr(.x = pop_2002_10 %>% dplyr::select(-PLAN_NO),
                       .y = pop_2011_19 %>% dplyr::select(-PLAN_NO),
                       ~ case_when(
                         is.nan(.x) & is.nan(.y)       ~ "not surveyed",
                         is.nan(.x) & .y == 0          ~ "uncolonised",
                         .x == 0    & is.nan(.y)       ~ "no 2nd mean",
                         .x == 0    & .y == 0          ~ "uncolonised",
                         is.nan(.x) & .y > 0           ~ "colonised",
                         .x == 0    & .y > 0           ~ "colonised",
                         .x!= 0     & .x < .y          ~ "increased",
                         .x > 0     & .y == 0          ~ "extinct",
                         .x > .y    & .y !=0           ~ "decreased",
                         .x == .y   & .x >0            ~ "stable",
                         .x > 0     & is.nan(.y)       ~ "no 2nd mean")) %>%
  cbind(PLAN_NO = pop_2011_19$PLAN_NO)

# Join population change data with sf OS squares inorder to plot
pop_chg_sf <- pop_chg_df %>% pivot_longer(cols = c(CU:SN), names_to = "Species", values_to = "Status") %>%
  left_join(sbbs_sq_sfc, by = "PLAN_NO") %>%
  st_as_sf(sf_column_name = "geometry")
```

Before any detailed statistical modelling was undertaken a simple analysis into how the population changed in each surveyed 1km square between 2002-2011 and 2012-2019. The 1 km squares shown (n=139) were those surveyed in both periods and where farmland waders colonized, increased, remained stable, declined or went extinct. This gave an initial view as to potential population trends between the two stated periods. Figure \@ref(fig:popStatusChg) shows the state changes between the two analysis periods.

```{r popStatusChg, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Population status change per Shetland BBS square - between 2002-10 and 2011-19' ,cache=TRUE}
pop_chg_sf %>% 
  ggplot(aes(x=Status)) + 
  geom_bar(aes(fill = Species), position = "dodge") +
  coord_flip() +
  theme_bw()
```
Figure \@ref(fig:aggPopChg) below shows an aggregation of certain categories; whereby extinct and decreased are grouped, and colonised and increased are grouped.

```{r aggPopChg, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Aggregate population status change per Shetland BBS square - between 2002-10 and 2011-19', cache=TRUE}
grouped_pop_chg_df <- pop_chg_sf %>% mutate(Status = case_when(
  Status == "extinct"   ~ "decline",
  Status == "decreased" ~ "decline",
  Status == "colonised" ~ "increase",
  Status == "increased" ~ "increase",
  TRUE                  ~ "no trend")) %>%
  st_drop_geometry()

grouped_pop_chg_df %>% 
  filter(Status == "decline" | Status == "increase") %>%
  ggplot(aes(x=Status)) + 
  geom_bar(aes(fill = Species), position = "dodge") +
  coord_flip() +
  theme_bw()
```
### Survey  Bootstrap

Shetland BBS volunteers were able to choose which squares they surveyed. The survey squares are therefore not randomly allocated across the Shetland archipelago. As a result of this non-randomised allocation there could be potential bias in the habitat types surveyed; for example, in-bye is closer to roads and housing than upland habitats. To test this a bootstrap of percentage cover of EUNIS habitat categories D, E and F (see \@ref(tab:eunisTable)) across all OS 1km squares was undertaken, and then compared to a bootstrap of the same data, but only those OS squares surveyed by volunteers as part of the Shetland BBS.


```{r bootstrap, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', cache = TRUE, fig.cap='Mean % cover per 1km$^2$ of EUNIS habitat types D, E and F, bootstrap sample of OSGB squares v boostrap of surveyed squares. R=1000'}
# load covar data for analysis
covars_df <- readRDS("~/Documents/GitHub/shetlandwaders/data/covars_df") %>%
  dplyr::select(PLAN_NO, D,E,F) %>%
  st_drop_geometry()

# determine SBBS sqaures
surveyed_PLAN_NO <- survey_BBS_2002_2018 %>% 
  distinct(.keep_all = T) %>%
  dplyr::select(PLAN_NO)

bbs_squares <- surveyed_PLAN_NO %>% 
  dplyr::left_join(covars_df,by="PLAN_NO") %>%
  na.omit() %>%
  st_drop_geometry()

# load bootstrap package
library(boot)

# function to generate bootstrap summary stats
summary_stats <- function(df, i) {
  sample = df[i, ]
  result <- sample %>% sapply(function(x) c(
    # standard deviation
    sd(x), 
    # mean
    mean(x),
    # standard error of mean
    sd(x)/sqrt(length(x))))
  
  return(result)
}

# Run bootstrap across all habitat across Shetland
b_allShet = boot(
  # dataset for bootstrap
  covars_df %>% dplyr::select(-PLAN_NO), 
  # function for sample
  summary_stats, 
  # Number of samples
  R=100)

# Run bootstrap on SBBS surveyed squares only
b_SBBS = boot(
  # dataset for bootstrap
  bbs_squares %>% dplyr::select(-PLAN_NO), 
  # function for sample
  summary_stats, 
  # Number of samples
  R=100)

library(data.table)
# Function to structure data
df_build <- function(df, sc) {
  df <- df %>%
    # Transpose dataframe
    t() %>% as.data.frame()
    # Add column names
    colnames(df) <- c("sd", "mean", "sde") 
    # Add habitat column from index
    df <- setDT(df, keep.rownames = "habitat")[]
    # Add data source column
    df$source <- as.factor(sc)
    return(df)
}

# Generate dataframe for plot
plot_df <- rbind(
  df_build(b_allShet$t0, "All OS squares"),
  df_build(b_SBBS$t0, "Survey only"))

# Plot 2018 data against bootstrap
# Error bars represent standard error of the mean
plot_df %>%
  ggplot(aes(x=habitat, y=mean, fill=source)) + 
    geom_bar(position=position_dodge(), stat="identity") +
    geom_errorbar(aes(ymin=mean-sde, ymax=mean+sde),
                  width=.2,                    # Width of the error bars
                  position=position_dodge(.9)) + 
  theme_bw()


```

This shows that grassland and heathland are significantly oversampled within the Shetland BBS surveys.