## Breeding wader abundance response to environmental covariates {#abun-resp_gam}

Generalised Additive Models (GAMs) for breeding wader abundance response across the five different species were generated from 2002–10 and 2011-19. A GAM was produced for each of the 10 environmental covariates (predictor variables), and for each of the two analysis periods.

```{r envCovarResponse, message=FALSE, warning=FALSE, echo=FALSE}
# Species detectability data
detect_df <- data.frame(
  # Waders under study
  Species = c("OC", "L", "RK", "CU", "SN"),
  d = c(0.803, # oystercatcher 
        0.723, # lapwing
        0.667, # redshank
        0.831, # curlew
        0.723 # snipe
        )) 

# Function to map a vector of years to a year range
period.cat <- function(x, lower, upper, by) {

 labs <- c(paste(seq(lower, upper - by, by = by),
                 seq(lower + by - 1, upper - 1, by = by),
                 sep = "-"),
           paste(upper-by + 1, upper, sep = "-"))

 cut(floor(x), 
     breaks = c(seq(lower, upper+1, by = by)),
     right = FALSE, labels=labs)
}

# Lower bound for year period
lw_yr = 2002
# Upper bound for year period
up_yr = 2019
# Number of years per period
yr_per_pd = 9

# Calculate total count across given period
#, by species and BBS square
waders_df <- bbs_df %>%
  mutate(period = period.cat(
    # Vector of years
    Year, lw_yr, up_yr, yr_per_pd )) %>%
  # Group data into rows that we want to sum by
  group_by(Species, period, PLAN_NO) %>%
  # Create a column for the mean observation count
  summarise(count_sm = sum(Count, na.rm=T)) %>%
  # Remove any row where there has been no survey
  # These are marked as NaN
  filter(!is.nan(count_sm))

# Generate df with 2011-19 as count, 2002-10 as offset
pop_chg_df <- waders_df %>%
  # Pivot wider to create two count columns per species and square: 2002-2010 and 2011-2019
  pivot_wider(names_from = period, values_from = count_sm) %>%
  # Only select surveys that were survyed in both periods, to prevent ln blow up! 
  filter(`2002-2010` > 0 & `2011-2019` >=0) %>%
  # Add marker to records
  mutate(period = "2002-2019") %>%
  # Change column names
  rename(count_sm = `2011-2019`,
         count_offset = `2002-2010`)

# Inorder to bind waders_df and pop_chg_df, an extra column
# count_offset must be added to waders_df
waders_df <- waders_df %>% mutate(count_offset = 0)
# Now can bind the two df's together into one df
waders_df <- rbind(waders_df, pop_chg_df)

# Create Waders data from for all observation data
wader_covars_df <- left_join(waders_df, histo_covars, by = "PLAN_NO") %>%
  # Join wader detectability values into main dataset
  left_join(.,detect_df, by="Species")
```


```{r fitGAM, message=FALSE, warning=FALSE, echo=FALSE}
# load gam library
library(mgcv)

wader_df <- wader_covars_df %>% 
  # Structure data by covariate and then group
    pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                     'Bare peat % cover':'Improved grassland % cover'), 
                 names_to = "covariate",
                 values_to = "value") %>%
  group_by(period, Species, covariate) %>%
  nest() %>%
  mutate(model = map2(.x = data,
                      .y = period,
                      # Generate a univariate GAM model for wader count against covariate (value).
                      # Repeat for each combination of wader and covariate 
                      ~ gam(count_sm ~ s(X,Y, k=20) + value,
                                      # Apply a different offset depending on whether this gam
                                      # models population change or only covariate response
                                      offset = case_when(.y == "2002-2019" ~ log(count_offset), 
                                                         .y != "2002-2019" ~ log(d) + log(9)),
                                      family = poisson(link="log"),
                                      method = "REML",
                                      data = .x)),
    # Summarise model config for reporting
    model_smry = map(model, ~tidy(.,parametric=TRUE))) %>%
  unnest(cols=model_smry) %>%
  filter(term != "(Intercept)") %>%
  dplyr::select(-term)

```

Appendix \@ref(gam-abun-response-params) details the GAM parameters for the model fitting process, for the two periods where abundance response (density) was modelled, and the associated plots showing breeding wader density against each environmental covariate are shown in Appendix \@ref(gam-abun-response-plots). The statistically significant correlations between breeding wader density and environmental covariates are summarised for each species in heatmaps; Figure \@ref(fig:heatMap200210) for 2002-10 and Figure \@ref(fig:heatMap201119) for 2011-2019. 

```{r heatMap200210, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of associations between breeding wader density and environmental covariate, between 2002 and 2010 inclusive'}
covar_species_assoc_2002_10 <- data.frame(
  Covariate = c('Elevation (m)',
          'Bog % cover', 
          'Grassland % cover', 
          'Heathland % cover', 
          'pH', 
          'Topsoil Organic Carbon %', 
          'Available Water Capacity (kPA)',
          'Distance to sea (m)', 
          'Bare peat % cover', 
          'Improved grassland % cover'),
  CU = c("Neg","Neg","Pos","NS","NS","Neg","Neg","Pos","Neg","Pos"),
  L =  c("Neg","Neg","Pos","NS","NS","Neg","Neg","Neg","NS","NS"),
  OC = c("Neg","Neg","Pos","Pos","NS","Neg","Neg","Pos","NS","NS"),
  RK = c("Neg","Neg","Pos","NS","NS","Neg","Neg","NS","Neg","Pos"),
  SN = c("NS", "NS","Pos","Pos","Pos","Neg","NS","Pos","Neg","Neg")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

covar_species_assoc_2002_10 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("blue", "grey", "red")) +
  theme_minimal()
```
For the 2002-10 survey period in Figure \@ref(fig:heatMap200210), it can be seen that pH is only statistically significant for one species (Snipe), whilst topsoil organic carbon and grassland are oppositely correlated. Distance to sea is perhaps the most interesting covariate in that Lapwing show greater density at the coast whilst Curlew, Oystercatcher and Snipe show greater densities inland. 

```{r heatMap201119, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of associations between breeding wader density and environmental covariate, between 2011 and 2019 inclusive'}
covar_species_assoc_2011_19 <- data.frame(
  Covariate = c('Elevation (m)',
                'Bog % cover', 
                'Grassland % cover', 
                'Heathland % cover', 
                'pH', 
                'Topsoil Organic Carbon %', 
                'Available Water Capacity (kPA)',
                'Distance to sea (m)', 
                'Bare peat % cover', 
                'Improved grassland % cover'),
  
  CU = c("Neg","Neg","Pos","Pos","Pos","Neg","Neg","Pos","Neg","Pos"),
  L =  c("Neg","Neg","Pos","Pos","NS","Neg","Neg","Neg","Neg","Pos"),
  OC = c("Neg","Neg","Pos","Pos","Pos","Neg","Neg","Neg","Neg","Pos"),
  RK = c("Neg","Neg","Pos","Pos","NS","Neg","Neg","NS","Neg","Pos"),
  SN = c("NS", "Neg","Pos","Pos","NS","NS","Pos","Pos","Neg","Neg")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

covar_species_assoc_2011_19 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("blue", "red","grey")) +
  theme_minimal()
```

In Figure \@ref(fig:heatMap201119) we can see that there are fewer associations that are not statistically significant. For Curlew it is seen that all covariate associations are now statistically significant (versus 2002-2010 where pH and Heathland cover were not significant). Oystercatcher also have all covariates with statistically significant associations, but for the later survey period the distance to association is now negative with increase distance. Lapwing only have one covariate, pH, that is not statistically significant. For Redshank the main change in the later survey period is that Heathland percentage cover is now statistically significant, and has a positive association. Snipe now have a positive association with available water capacity and a negative association with percentage bog cover.

## Population change response against environmental covariates

A third model was generated by using the abundance response from the first two GAMs in \@ref(abun-resp_gam) to generate the ratio of population change over the two response periods. By using the response of the 2002-2010 wader densities as the offset for the 2011-19 densities, a third series of GAMs were fitted to show the ratio of population change in response to environmental covariates. This is shown in Figure \@ref(fig:heatMapPopChg).

```{r heatMapPopChg, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of population change ratio associations between breeding wader density and environmental covariate, between 2002 and 2019 inclusive'}
pop_chg_2002_19 <- data.frame(
  Covariate = c('Elevation (m)',
                'Bog % cover', 
                'Grassland % cover', 
                'Heathland % cover', 
                'pH', 
                'Topsoil Organic Carbon %', 
                'Available Water Capacity (kPA)',
                'Distance to sea (m)', 
                'Bare peat % cover', 
                'Improved grassland % cover'),
  
  CU = c("Neg","Neg","NS","Pos","NS","NS","NS","Neg","NS","NS"),
  L =  c("NS","NS","NS","Pos","Pos","Neg","Neg","NS","NS","Pos"),
  OC = c("NS","NS","NS","NS","Pos","NS","Pos","Neg","NS","Pos"),
  RK = c("Pos","Pos","Neg","Pos","NS","NS","Pos","NS","NS","NS"),
  SN = c("NS", "NS","NS","NS","Neg","Pos","Pos","NS","NS","NS")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

pop_chg_2002_19 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("blue","grey","red")) +
  theme_minimal()
```

Figure \@ref(fig:heatMapPopChg) suggests that the environmental covariates with that have had the most positive associations with breeding wader density are heathland percentage cover and available water capacity, whilst the percentage of bare peatland has had no statistical significance, followed by the percentage grassland cover that has only one negative association with the population change ratio of Redshank.

The model parameters and associated plots for population change ratio modelling are shown in Appendix \@ref(gam-pop-chg-params) and Appendix \@ref(gam-pop-chg-plots) respectively.

