## Environmental covariate response

GAMs for breeding wader abundance response across the five different species were generated from 2002–10 and 2011-19. They included 10 environmental predictor variables (covariates), and the model parameters for all species across the two response periods can be seen in Table \@ref(tab:tableGamParamsAbundance). A third model was generated by taking the abundance response from the first two models to generate the ratio of population change between the two response periods.

```{r envCovarResponse, message=FALSE, warning=FALSE, echo=FALSE}
# Species detectability data
detect_df <- data.frame(
  # Waders under study
  Species = c("OC", "L", "RK", "CU", "SN"),
  d = c(0.803, # oystercatcher 
        0.723, # lapwing
        0.667, # redshank
        0.831, # curlew
        0.723 # snipe
        )) 

# Function to map a vector of years to a year range
period.cat <- function(x, lower, upper, by) {

 labs <- c(paste(seq(lower, upper - by, by = by),
                 seq(lower + by - 1, upper - 1, by = by),
                 sep = "-"),
           paste(upper-by + 1, upper, sep = "-"))

 cut(floor(x), 
     breaks = c(seq(lower, upper+1, by = by)),
     right = FALSE, labels=labs)
}

# Lower bound for year period
lw_yr = 2002
# Upper bound for year period
up_yr = 2019
# Number of years per period
yr_per_pd = 9

# Calculate total count across given period
#, by species and BBS square
waders_df <- bbs_df %>%
  mutate(period = period.cat(
    # Vector of years
    Year, lw_yr, up_yr, yr_per_pd )) %>%
  # Group data into rows that we want to sum by
  group_by(Species, period, PLAN_NO) %>%
  # Create a column for the mean observation count
  summarise(count_sm = sum(Count, na.rm=T)) %>%
  # Remove any row where there has been no survey
  # These are marked as NaN
  filter(!is.nan(count_sm))

# Generate df with 2011-19 as count, 2002-10 as offset
pop_chg_df <- waders_df %>%
  # Pivot wider to create two count columns per species and square: 2002-2010 and 2011-2019
  pivot_wider(names_from = period, values_from = count_sm) %>%
  # Only select surveys that were survyed in both periods, to prevent ln blow up! 
  filter(`2002-2010` > 0 & `2011-2019` >=0) %>%
  # Add marker to records
  mutate(period = "2002-2019") %>%
  # Change column names
  rename(count_sm = `2011-2019`,
         count_offset = `2002-2010`)

# Inorder to bind waders_df and pop_chg_df, an extra column
# count_offset must be added to waders_df
waders_df <- waders_df %>% mutate(count_offset = 0)
# Now can bind the two df's together into one df
waders_df <- rbind(waders_df, pop_chg_df)

# Create Waders data from for all observation data
wader_covars_df <- left_join(waders_df, histo_covars, by = "PLAN_NO") %>%
  # Join wader detectability values into main dataset
  left_join(.,detect_df, by="Species")
```


```{r fitGAM, message=FALSE, warning=FALSE, echo=FALSE}
# load gam library
library(mgcv)

wader_df <- wader_covars_df %>% 
  # Structure data by covariate and then group
    pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                     'Bare peat % cover':'Improved grassland % cover'), 
                 names_to = "covariate",
                 values_to = "value") %>%
  group_by(period, Species, covariate) %>%
  nest() %>%
  mutate(model = map2(.x = data,
                      .y = period,
                      # Generate a univariate GAM model for wader count against covariate (value).
                      # Repeat for each combination of wader and covariate 
                      ~ gam(count_sm ~ s(X,Y, k=20) + value,
                                      # Apply a different offset depending on whether this gam
                                      # models population change or only covariate response
                                      offset = case_when(.y == "2002-2019" ~ log(count_offset), 
                                                         .y != "2002-2019" ~ log(d) + log(9)),
                                      family = poisson(link="log"),
                                      method = "REML",
                                      data = .x)),
    # Summarise model config for reporting
    model_smry = map(model, ~tidy(.,parametric=TRUE))) %>%
  unnest(cols=model_smry) %>%
  filter(term != "(Intercept)") %>%
  dplyr::select(-term)

```

### Environmental covariate GAM model parameters

Table \@ref(tab:tableGamParamsAbundance) shows the GAM model parameters for the model fits for the two periods where abundance response (density) was modelled, and the associated plots showing breeding wader density against each environmental covariate are shown in Figure \@ref(fig:responsePlots). The statistically significant correlations between breeding wader density and environmental covariate are summarised for each species in heatmaps; Figure \@ref(fig:heatMap200210) for 2002-10 and Figure \@ref(fig:heatMap201119) for 2011-2019. 

```{r heatMap200210, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of associations between breeding wader density and environmental covariate, between 2002 and 2010 inclusive'}
covar_species_assoc_2002_10 <- data.frame(
  Covariate = c('Elevation (m)',
          'Bog % cover', 
          'Grassland % cover', 
          'Heathland % cover', 
          'pH', 
          'Topsoil Organic Carbon %', 
          'Available Water Capacity (kPA)',
          'Distance to sea (m)', 
          'Bare peat % cover', 
          'Improved grassland % cover'),
  CU = c("Neg","Neg","Pos","NS","NS","Neg","Neg","Pos","Neg","Pos"),
  L =  c("Neg","Neg","Pos","NS","NS","Neg","Neg","Neg","NS","NS"),
  OC = c("Neg","Neg","Pos","Pos","NS","Neg","Neg","Pos","NS","NS"),
  RK = c("Neg","Neg","Pos","NS","NS","Neg","Neg","NS","Neg","Pos"),
  SN = c("NS", "NS","Pos","Pos","Pos","Neg","NS","Pos","Neg","Neg")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

covar_species_assoc_2002_10 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("blue", "grey", "red")) +
  theme_minimal()
```

For the 2002-10 survey period in Figure \@ref(fig:heatMap200210), it can be seen that pH is only statistically significant for one species (Snipe), whilst topsoil organic carbon and grassland are oppositely correlated. Distance to sea is perhaps the most interesting covariate in that Lapwing show greater density at the coast whilst Curlew, Oystercatcher and Snipe show greater densities inland. 

```{r heatMap201119, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of associations between breeding wader density and environmental covariate, between 2011 and 2019 inclusive'}
covar_species_assoc_2011_19 <- data.frame(
  Covariate = c('Elevation (m)',
                'Bog % cover', 
                'Grassland % cover', 
                'Heathland % cover', 
                'pH', 
                'Topsoil Organic Carbon %', 
                'Available Water Capacity (kPA)',
                'Distance to sea (m)', 
                'Bare peat % cover', 
                'Improved grassland % cover'),
  
  CU = c("Neg","Neg","Pos","Pos","Pos","Neg","Neg","Pos","Neg","Pos"),
  L =  c("Neg","Neg","Pos","Pos","NS","Neg","Neg","Neg","Neg","Pos"),
  OC = c("Neg","Neg","Pos","Pos","Pos","Neg","Neg","Neg","Neg","Pos"),
  RK = c("Neg","Neg","Pos","Pos","NS","Neg","Neg","NS","Neg","Pos"),
  SN = c("NS", "Neg","Pos","Pos","NS","NS","Pos","Pos","Neg","Neg")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

covar_species_assoc_2011_19 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("blue", "red","grey")) +
  theme_minimal()
```

In Figure \@ref(fig:heatMap201119) we can see that there are fewer associations that are not statistically significant. For Curlew it is seen that all covariate associations are now statistically significant (versus 2002-2010 where pH and Heathland cover were not significant). Oystercatcher also have all covariates with statistically significant associations, but for the later survey period the distance to association is now negative with increase distance. Lapwing only have one covariate, pH, that is not statistically significant. For Redshank the main change in the later survey period is that Heathland percentage cover is now statistically significant, and has a positive association. Snipe now have a positive association with available water capacity and a negative association with percentage bog cover.

```{r tableGamParamsAbundance, message=FALSE, warning=FALSE, echo=FALSE, fig.align='left', fig.asp= 1.1, fig.cap='GAM parameter estimates, standard errors, z and P values from models incorporating parameteric terms associated with abundance (in 2002-2010 and 2011-2019)'}
# create table of results of gam fit for each species - abundance response only
library(kableExtra)
wader_df %>%
    dplyr::select(Species, period,covariate,estimate:p.value) %>%
    filter(period != "2002-2019") %>%
    kableExtra::kable(
      align = c('c','l','l','c','c','c','c'),
      col.names = c("Species","Response","Covariate",	"Estimate",	"se", "z", "p-value")) %>%
    kable_styling(full_width = F, font_size = 10) %>%
    column_spec(1, bold = T) %>%
    collapse_rows(columns = 1:2, valign = "top") 
```

### Environmental covariate GAM model plots 

Figures \@ref(fig:responsePlots)show the response of wader density to each environmental covariate, for the survey periods 2002-10 and 2011-19.

```{r responsePlots, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Plots show abundance response to environmental covariates for a GAM'}
# Generate prediction data so that we can plot parametric terms
wader_df <- wader_df %>% 
  mutate(model_pred = map2(
    .x = model, 
    .y = covariate, 
    # A function to generate a response with parametric covariate only
    ~tidymv::get_gam_predictions(model = .x,
                                 series = .y, 
                                 # We want the response not link
                                 transform = exp)))

# Generate plot for each model
wader_df <- wader_df %>%
  # Use a column to store plots
  mutate(
    plots = pmap(
      list(w = Species , c = covariate, cs = p.value, m = model_pred, d = period),
      function(w,c,cs,m,d){
         ggplot(data = m) +
         # Plot parametric coefficient - red for plots with p <= 0.05
         geom_line(aes(x=value,y=count_sm), colour = if_else(cs>0.05, 1,2)) + 
         # Plot confidence interval
         geom_ribbon(aes(x =value, ymin=CI_lower,ymax=CI_upper), alpha=0.2) +
         # x axis label set to covariate in model  
         xlab(c) +
         # y axis label set to wader name  
         ylab(paste(w, if_else(d == "2002-2019", "pop chg ratio", "density"), sep = " ")) +
         theme_bw(base_size = 6) +
         theme(legend.position = 'none')
      }))

# Library to arrange plots into pages
library(gridExtra)

# Generate output
wader_df %>% 
  # We dont want the population change plots
  filter(period !="2002-2019") %>%
  ungroup %>% 
  dplyr::select(period, plots) %>%
  # Group plots into date ranges
  group_by(period) %>%
  # Map plotting over group
  group_map(
    ~marrangeGrob(.x$plots, 
                  nrow =5, 
                  ncol = 2,
                  # .y gives the key of the grouping variable
                  top = as.character(.y)
                  ))
```

## Population change model against environmental covariates

By using the response of the 2002-2010 wader densities as the offset for the 2011-19 densities, a third series of GAMs were fitted to show the ratio of population change in response to environmental covariates. This is shown in Figure \@ref(fig:heatMapPopChg).

```{r heatMapPopChg, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of population change ratio associations between breeding wader density and environmental covariate, between 2002 and 2019 inclusive'}
pop_chg_2002_19 <- data.frame(
  Covariate = c('Elevation (m)',
                'Bog % cover', 
                'Grassland % cover', 
                'Heathland % cover', 
                'pH', 
                'Topsoil Organic Carbon %', 
                'Available Water Capacity (kPA)',
                'Distance to sea (m)', 
                'Bare peat % cover', 
                'Improved grassland % cover'),
  
  CU = c("Neg","Neg","NS","Pos","NS","NS","NS","Neg","NS","NS"),
  L =  c("NS","NS","NS","Pos","Pos","Neg","Neg","NS","NS","Pos"),
  OC = c("NS","NS","NS","NS","Pos","NS","Pos","Neg","NS","Pos"),
  RK = c("Pos","Pos","Neg","Pos","NS","NS","Pos","NS","NS","NS"),
  SN = c("NS", "NS","NS","NS","Neg","Pos","Pos","NS","NS","NS")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

pop_chg_2002_19 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("blue","grey","red")) +
  theme_minimal()
```

Figure \@ref(fig:heatMapPopChg) suggests that the environmental covariates with that have had the most positive associations with breeding wader density are heathland percentage cover and available water capacity, whilst the percentage of bare peatland has had no statistical significance, followed by the percentage grassland cover that has only one negative association with the population change ratio of Redshank.

The model parameters and associated plots for population change ratio modelling are shown in Figures \@ref(popChgPlots) and Table \@ref(tab:tableGamParamsPopChg) respectively.

```{r tableGamParamsPopChg, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='GAM parameter estimates, standard errors, z and P values from models incorporating parameteric terms associated with abundance (in 2002-2010 and 2011-2019)'}
# create table of results of gam fit for each species - abundance response only
wader_df %>%
    dplyr::select(Species, period, covariate,estimate:p.value) %>%
    filter(period == "2002-2019") %>%
    kableExtra::kable(
      align = c('c','l','l','c','c','c','c'),
      col.names = c("Species","Period", "Covariate","Estimate",	"se", "z", "p-value")) %>%
    kable_styling(full_width = F, font_size = 10) %>%
    column_spec(1, bold = T) %>%
    collapse_rows(columns = 1:2, valign = "top") 
```


```{r popChgPlots, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Plots show population change to environmental covariates for a GAM, acrossall wader species'}
# Generate output
wader_df %>% 
  # We only want the population change plots
  filter(period == "2002-2019") %>%
  ungroup %>% 
  dplyr::select(period, plots) %>%
  # Group plots into date ranges
  group_by(period) %>%
  # Map plotting over group
  group_map(
    ~marrangeGrob(.x$plots, 
                  nrow =5, 
                  ncol = 2,
                  # .y gives the key of the grouping variable
                  top = as.character(.y)
                  ))
```