## Environmental covariate response

```{r envCovarResponse, message=FALSE, warning=FALSE, echo=FALSE}
# Species detectability data
detect_df <- data.frame(
  # Waders under study
  Species = c("OC", "L", "RK", "CU", "SN"),
  d = c(0.803, # oystercatcher 
        0.723, # lapwing
        0.667, # redshank
        0.831, # curlew
        0.723 # snipe
        )) 

# Function to map a vector of years to a year range
period.cat <- function(x, lower, upper, by) {

 labs <- c(paste(seq(lower, upper - by, by = by),
                 seq(lower + by - 1, upper - 1, by = by),
                 sep = "-"),
           paste(upper-by + 1, upper, sep = "-"))

 cut(floor(x), 
     breaks = c(seq(lower, upper+1, by = by)),
     right = FALSE, labels=labs)
}

# Lower bound for year period
lw_yr = 2002
# Upper bound for year period
up_yr = 2019
# Number of years per period
yr_per_pd = 9

# Calculate total count across given period
#, by species and BBS square
waders_df <- bbs_df %>%
  mutate(period = period.cat(
    # Vector of years
    Year, lw_yr, up_yr, yr_per_pd )) %>%
  # Group data into rows that we want to sum by
  group_by(Species, period, PLAN_NO) %>%
  # Create a column for the mean observation count
  summarise(count_sm = sum(Count, na.rm=T)) %>%
  # Remove any row where there has been no survey
  # These are marked as NaN
  filter(!is.nan(count_sm))

# Generate df with 2011-19 as count, 2002-10 as offset
pop_chg_df <- waders_df %>%
  # Pivot wider to create two count columns per species and square: 2002-2010 and 2011-2019
  pivot_wider(names_from = period, values_from = count_sm) %>%
  # Only select surveys that were survyed in both periods, to prevent ln blow up! 
  filter(`2002-2010` > 0 & `2011-2019` >=0) %>%
  # Add marker to records
  mutate(period = "2002-2019") %>%
  # Change column names
  rename(count_sm = `2011-2019`,
         count_offset = `2002-2010`)

# Inorder to bind waders_df and pop_chg_df, an extra column
# count_offset must be added to waders_df
waders_df <- waders_df %>% mutate(count_offset = 0)
# Now can bind the two df's together into one df
waders_df <- rbind(waders_df, pop_chg_df)

# Create Waders data from for all observation data
wader_covars_df <- left_join(waders_df, histo_covars, by = "PLAN_NO") %>%
  # Join wader detectability values into main dataset
  left_join(.,detect_df, by="Species")
```


```{r fitGAM, message=FALSE, warning=FALSE, echo=FALSE}
# load gam library
library(mgcv)

wader_df <- wader_covars_df %>% 
  # Structure data by covariate and then group
    pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                     'Bare peat % cover':'Improved grassland % cover'), 
                 names_to = "covariate",
                 values_to = "value") %>%
  group_by(period, Species, covariate) %>%
  nest() %>%
  mutate(model = map2(.x = data,
                      .y = period,
                      # Generate a univariate GAM model for wader count against covariate (value).
                      # Repeat for each combination of wader and covariate 
                      ~ gam(count_sm ~ s(X,Y, k=20) + value,
                                      # Apply a different offset depending on whether this gam
                                      # models population change or only covariate response
                                      offset = case_when(.y == "2002-2019" ~ log(count_offset), 
                                                         .y != "2002-2019" ~ log(d) + log(9)),
                                      family = poisson(link="log"),
                                      method = "REML",
                                      data = .x)),
    # Summarise model config for reporting
    model_smry = map(model, ~tidy(.,parametric=TRUE))) %>%
  unnest(cols=model_smry) %>%
  filter(term != "(Intercept)") %>%
  dplyr::select(-term)

```

### Environmental covariate GAM model parameters

Table [ref] shows the GAM model parameters for the model fits for the two periods were abundance response was modelled.

```{r tableGamParamsAbundance, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='GAM parameter estimates, standard errors, z and P values from models incorporating parameteric terms associated with abundance (in 2002-2010 and 2011-2019)'}
# create table of results of gam fit for each species - abundance response only
library(kableExtra)
wader_df %>%
    dplyr::select(Species, period,covariate,estimate:p.value) %>%
    filter(period != "2002-2019") %>%
    kableExtra::kable(
      align = c('c','l','l','c','c','c','c'),
      col.names = c("Species","Response","Covariate",	"Estimate",	"se", "z", "p-value")) %>%
    kable_styling(full_width = F, font_size = 10) %>%
    column_spec(1, bold = T) %>%
    collapse_rows(columns = 1:2, valign = "top") 
```
### Environmental covariate GAM model plots 

Given the prediction a plot can be generated to show the response of wader density to each covariate model.

```{r responsePlots, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Plots show abundance response to environmental covariates for a GAM'}
# Generate prediction data so that we can plot parametric terms
wader_df <- wader_df %>% 
  mutate(model_pred = map2(
    .x = model, 
    .y = covariate, 
    # A function to generate a response with parametric covariate only
    ~tidymv::get_gam_predictions(model = .x,
                                 series = .y, 
                                 # We want the response not link
                                 transform = exp)))

# Generate plot for each model
wader_df <- wader_df %>%
  # Use a column to store plots
  mutate(
    plots = pmap(
      list(w = Species , c = covariate, cs = p.value, m = model_pred, d = period),
      function(w,c,cs,m,d){
         ggplot(data = m) +
         # Plot parametric coefficient - red for plots with p <= 0.05
         geom_line(aes(x=value,y=count_sm), colour = if_else(cs>0.05, 1,2)) + 
         # Plot confidence interval
         geom_ribbon(aes(x =value, ymin=CI_lower,ymax=CI_upper), alpha=0.2) +
         # x axis label set to covariate in model  
         xlab(c) +
         # y axis label set to wader name  
         ylab(paste(w, if_else(d == "2002-2019", "pop chg ratio", "density"), sep = " ")) +
         theme_bw(base_size = 6) +
         theme(legend.position = 'none')
      }))

# Library to arrange plots into pages
library(gridExtra)

# Generate output
wader_df %>% 
  # We dont want the population change plots
  filter(period !="2002-2019") %>%
  ungroup %>% 
  dplyr::select(period, plots) %>%
  # Group plots into date ranges
  group_by(period) %>%
  # Map plotting over group
  group_map(
    ~marrangeGrob(.x$plots, 
                  nrow =5, 
                  ncol = 2,
                  # .y gives the key of the grouping variable
                  top = as.character(.y)
                  ))
```


## Population change model against environmental covariates

```{r tableGamParamsPopChg, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='GAM parameter estimates, standard errors, z and P values from models incorporating parameteric terms associated with abundance (in 2002-2010 and 2011-2019)'}
# create table of results of gam fit for each species - abundance response only
wader_df %>%
    dplyr::select(Species, period, covariate,estimate:p.value) %>%
    filter(period == "2002-2019") %>%
    kableExtra::kable(
      align = c('c','l','l','c','c','c','c'),
      col.names = c("Species","Period", "Covariate","Estimate",	"se", "z", "p-value")) %>%
    kable_styling(full_width = F, font_size = 10) %>%
    column_spec(1, bold = T) %>%
    collapse_rows(columns = 1:2, valign = "top") 
```

Plots for the population change model

```{r popChgPlots, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Plots show population change to environmental covariates for a GAM, acrossall wader species'}
# Generate output
wader_df %>% 
  # We only want the population change plots
  filter(period == "2002-2019") %>%
  ungroup %>% 
  dplyr::select(period, plots) %>%
  # Group plots into date ranges
  group_by(period) %>%
  # Map plotting over group
  group_map(
    ~marrangeGrob(.x$plots, 
                  nrow =5, 
                  ncol = 2,
                  # .y gives the key of the grouping variable
                  top = as.character(.y)
                  ))
```