## Environmental covariate analysis

Each of the covariates described in \ref was generated for each Shetland BBS squares (n=3992). 

### Histogram of environmental covariates 

Figure [ref] shows histograms of how these covariates are distributed across Shetland.

```{r covarHisto, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Histograms of environmental covariates across all of Shetland'}
# Load covar data for each BBS square
# Transform sf geom to X & Y
covars_df <- readRDS("~/Documents/GitHub/shetlandwaders/data/covars_df") %>%
  st_centroid() %>% 
  cbind(., sf::st_coordinates(.)) %>% 
  st_drop_geometry()

# Filter required covariates and name appropriately
histo_covars <- covars_df %>%
  # Rename covariates
  rename('Elevation (m)'                   = elev,
         'Bog % cover'                     = D,
         'Grassland % cover'               = E,
         'Heathland % cover'               = `F`,
         'pH'                              = wgt_pH,
         'Topsoil Organic Carbon %'        = wgt_orgCarbon,
         'Available Water Capacity (kPA)'  = AWC,
         'Distance to sea (m)'             = dist_sea,
         'Bare peat % cover'               = bare_pc_cvr,
         'Improved grassland % cover'      = imp_gr_pc_df,
         'Marginal entropy'                = ent,
         'Conditional entropy'             = condent,
         'Joint entropy'                   = joinent,
         'Mutual information'              = mutinf,
         'Relative mutual informaton'      = relmutinf)

# Plot histogram
histo_covars %>%
  # pivot long by species
  pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                     'Bare peat % cover':'Improved grassland % cover'), 
               names_to = "covar",
               values_to = "value") %>%
  # Remove zero values as this makes it hard to see the distribution
  filter(value >0) %>%
  ggplot(aes(x = value)) + 
  # plot the histogram
  stat_bin(color = "black", bins=40) +
  theme_minimal() +
  labs(x = "Environmental covariate", y = "Frequency") +
  facet_wrap(~covar, ncol=2, scales="free")

```

### Histogram of environmental covariates for Shetland BBS squares only

This can be contrasted with covariate histograms for only those OS squares (n=139) that were surveyed as part of the Shetland BBS.

```{r sbbsHisto, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Histograms of environmental covariates across only those squares surveyed as part of the Shetland BBS'}
# Load covar data for each BBS sq}

bbs_covars <- surveyed_PLAN_NO %>% 
  dplyr::left_join(histo_covars, by="PLAN_NO") %>%
  na.omit() %>%
  st_drop_geometry()

# Plot histogram
bbs_covars %>%
  # pivot long by species
  pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                      'Bare peat % cover':'Improved grassland % cover'), 
               names_to = "covar",
               values_to = "value") %>%
  # Remove zero values as this makes it hard to see the distribution
  filter(value >0) %>%
  ggplot(aes(x = value)) + 
  # plot the histogram
  stat_bin(color = "black") +
  theme_minimal() +
  labs(x = "Environmental covariate", y = "Frequency") +
  facet_wrap(~covar, ncol=2, scales="free")

```

### Density plots

Density plot of all breeding wader count data against environmental covariates are shown in figure [ref]

```{r densityPlot, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Density plots of environmental covariates against breeding wader count data'}

# Create joined dataset of wader counts and covariate data
wader_covars_df <- left_join(bbs_df, histo_covars, by="PLAN_NO" ) %>%
  na.omit

# Create density plots
wader_covars_df %>% 
  # pivot to show lc % for each wader recorded
  pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                      'Bare peat % cover':'Improved grassland % cover'), 
               names_to = "covar",
               values_to = "value") %>%
  # only take presence observations where lc percentage > 0%
  filter(value > 0 ) %>%
  # Select the data we want - value = lc %
  dplyr::select(covar, value) %>%
  group_by(covar) %>%
  # Generate a histogram
  ggplot(aes(x=value)) +
    geom_histogram(stat="bin", bins=35, aes(y=..density..)) +
    # Add density plot
    geom_density(alpha=.2, fill="#FF6666") +
    facet_wrap(~covar, ncol=2, scales="free") +
    theme_bw() +
    xlab("Percentage of habitat type (25% bins) across all surveyed 1km squares")

```