## Environmental covariate analysis

Each of the covariates described in Section \@ref(environmental-covariates) were generated for each OS 1km Shetland square (n=3992). 

### Histogram of environmental covariates 

Figure \@ref(fig:covarHisto) shows histograms for all environmental covariates across the Shetland archipelago, at 1km resolution. It can be seen that bog type habitat predominates across Shetland, and that majority of the landscape is at less than 100m elevation. The mode for the pH is around 4, which is typical for acidic peatland [@Paterson2011-ky]. The topsoil carbon content shows that the majority of the Shetland soils have a high organic carbon content; again this is typical of peatland [@Paterson2011-ky]. In contrast a typical mineral rich soil in southern England would have c.3-5% organic carbon content. Note that the percentage landcover of improved grassland is relatively low compared to the more general grassland classification. This shows that agricultural intensification is relatively low in Shetland.

```{r covarHisto, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp = 1.1, fig.cap='Histograms of environmental covariates across all of Shetland'}
# Load covar data for each BBS square
# Transform sf geom to X & Y
covars_df <- readRDS("~/Documents/GitHub/shetlandwaders/data/covars_df") %>%
  st_centroid() %>% 
  cbind(., sf::st_coordinates(.)) %>% 
  st_drop_geometry()

# Filter required covariates and name appropriately
histo_covars <- covars_df %>%
  # Rename covariates
  rename('Elevation (m)'                   = elev,
         'Bog % cover'                     = D,
         'Grassland % cover'               = E,
         'Heathland % cover'               = `F`,
         'pH'                              = wgt_pH,
         'Topsoil Organic Carbon %'        = wgt_orgCarbon,
         'Available Water Capacity (kPA)'  = AWC,
         'Distance to sea (m)'             = dist_sea,
         'Bare peat % cover'               = bare_pc_cvr,
         'Improved grassland % cover'      = imp_gr_pc_df,
         'Marginal entropy'                = ent,
         'Conditional entropy'             = condent,
         'Joint entropy'                   = joinent,
         'Mutual information'              = mutinf,
         'Relative mutual informaton'      = relmutinf)

# Plot histogram
histo_covars %>%
  # pivot long by species
  pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                     'Bare peat % cover':'Improved grassland % cover'), 
               names_to = "covar",
               values_to = "value") %>%
  # Remove zero values as this makes it hard to see the distribution
  filter(value >0) %>%
  ggplot(aes(x = value)) + 
  # plot the histogram
  stat_bin(color = "black", bins=40) +
  theme_minimal() +
  labs(x = "Environmental covariate", y = "Frequency") +
  facet_wrap(~covar, ncol=2, scales="free")

```

### Histogram of environmental covariates for Shetland BBS squares only {#bbs-env-cov}

This can be contrasted with covariate histograms for only those OS squares (n=139) that were surveyed as part of the Shetland BBS, as seen in Figure \@ref(fig:sbbsHisto). If these data are indicative of general breeding wader habitat preferences, it would seem that across all surveyed nesting wader species, there is a preference for wet but not water-logged habitat as seen in the AWC histogram. Also, the majority of breeding waders that were surveyed appear to nest within 1km of the coast. It appears that surveyed breeding waders also have a preference for grassland (both improved and unimproved) presents the majority of the habitat cover, over heathland.

```{r sbbsHisto, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Histograms of environmental covariates across only those squares surveyed as part of the Shetland BBS'}
# Load covar data for each BBS sq}

bbs_covars <- surveyed_PLAN_NO %>% 
  dplyr::left_join(histo_covars, by="PLAN_NO") %>%
  na.omit() %>%
  st_drop_geometry()

# Plot histogram
bbs_covars %>%
  # pivot long by species
  pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                      'Bare peat % cover':'Improved grassland % cover'), 
               names_to = "covar",
               values_to = "value") %>%
  # Remove zero values as this makes it hard to see the distribution
  filter(value >0) %>%
  ggplot(aes(x = value)) + 
  # plot the histogram
  stat_bin(color = "black") +
  theme_minimal() +
  labs(x = "Environmental covariate", y = "Frequency") +
  facet_wrap(~covar, ncol=2, scales="free")

```

### Density plots of environmental covariates

Density plots of all environmental covariates across Shetland BBS squares (n=139) are shown in Figure \@ref(fig:densityPlot). These figures provide an overlay to the histograms in Section \@ref(bbs-env-cov), and represent a smoothed version of a histogram to show the probability density function of the variable. Some distributions are highly skewed, such as distance to sea and elevation. Whilst other covariates like topsoil organic carbon and bog cover are largely a uniform distributed. None of the covariates appear to be normally distributed.

```{r densityPlot, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Density plots of environmental covariates against breeding wader count data'}

# Create joined dataset of wader counts and covariate data
wader_covars_df <- left_join(bbs_df, histo_covars, by="PLAN_NO" ) %>%
  na.omit

# Create density plots
wader_covars_df %>% 
  # pivot to show lc % for each wader recorded
  pivot_longer(cols=c('Elevation (m)':'Distance to sea (m)',
                      'Bare peat % cover':'Improved grassland % cover'), 
               names_to = "covar",
               values_to = "value") %>%
  # only take presence observations where lc percentage > 0%
  filter(value > 0 ) %>%
  # Select the data we want - value = lc %
  dplyr::select(covar, value) %>%
  group_by(covar) %>%
  # Generate a histogram
  ggplot(aes(x=value)) +
    geom_histogram(stat="bin", bins=35, aes(y=..density..)) +
    # Add density plot
    geom_density(alpha=.2, fill="#FF6666") +
    facet_wrap(~covar, ncol=2, scales="free") +
    theme_bw() +
    xlab("Percentage of habitat type (25% bins) across all surveyed 1km squares")

```