## Improved grassland connectivity

### Locate patches

```{r select_improved_patches}
# Find all improved grassland patches >= 2Ha
library(landscapemetrics)

# Crop raster to south mainland
south_ext <- c(433000, #xmin
               442173, #xmax 
               1106864,#ymin 
               1122554 #ymax
               ) 

crop_r <- crop(res_r, south_ext)

# Make crops and improved grassland that same value
crop_r <- reclassify(crop_r, cbind(5, 1))
# Locate all improved patches >= 2Ha
improved_patches <- crop_r %>% 
  lsm_p_area() %>% 
  filter(class == 1 & value >= 5) %>%
  pull(id)

# Get all points that are in improved patches >= 2Ha
patches_sf <- get_patches(crop_r, class = 1 )[[1]] %>% 
  rasterToPoints() %>% 
  as.data.frame() %>% 
  filter(layer %in% improved_patches) %>%
  sf::st_as_sf(coords = c(1,2), crs=27700) %>%
  rename(id = layer)

# Create convex hull over points within a patch
improved_polys_sf <- patches_sf %>% 
  nest(-id) %>%
  mutate(chull_poly = map_dfr(
    .x=data,
    ~{st_union(.x) %>% 
        st_convex_hull() %>% 
        st_as_sf(crs=27700)}
  )) %>%
  dplyr::select(-data)

# Plot results
improved_polys_sf$chull_poly %>% plot

# Plot improved polys over sentinel image
p <- ggplot() + ggRGB(img = sent,
                      r = 4,
                      g = 3,
                      b = 2,
                      stretch = 'lin', ggLayer = T) +
xlim(south_ext[1:2]) + ylim(south_ext[3:4]) +
geom_sf(alpha = 0.3, aes(col="improved"), data = improved_polys_sf$chull_poly$x) 
p
```

# Connectivity analysis

The hypothesis is that breeding sites with higher abundance are located close to a large area of nearby improved grassland. In particualr Curlew are known to walk their chicks from the breeding site to improved grassland, in order to feed.

First generate an adjacency matrix between breeding sites and improved grassland sites.

```{r adj_matrix}
# Load survey data
rspb_df <- readRDS("~/Documents/GitHub/shetlandwaders/data/rspb_covars_df")
# Pivot longer
rspb_df <- rspb_df %>% pivot_longer(cols=CU:SN, names_to = "Species", values_to = "Count")

# Select a study species
study_sp = "OC"
# Select site data for our study
species_sites_sf <- rspb_df %>% 
  filter(Species == study_sp) %>% 
  select(Name, Area_km2, Count, geometry) %>%
  st_as_sf() 
  #st_crop(xmin =  433497.1, xmax = 442282.1, ymin = 1106864 , ymax = 1122394, crs=27700) #%>%

# Site names
breeding_sites <- species_sites_sf$Name %>% unique()

# Generate matrix of pairwise shortest distances between breeding sites (cols) and improved grassland (rows)
# Define threshold distance - the distance beyond which, a wader will not travel for food during the breeding season
threshold <- 500

adj_matrix <- st_distance(
  improved_polys_sf$chull_poly, 
  species_sites_sf, 
  by_element = FALSE) %>% 
  mapply(FUN=as.numeric) %>%
  matrix(ncol=NROW(species_sites_sf), 
         nrow=NROW(improved_polys_sf)) %>% 
  as.data.frame()

# Decay function
weight_fn <- function(x,threshold){
  if (x > threshold) {
    return(0)
    }
  if (x <= threshold) {
      weight = (threshold - x)/(threshold + x)
      return(weight)
  }
}

# Calculate node strength
# This is the sum of weighted edges, where
# each edge is weighted by a decay function
node_strength <- adj_matrix %>% 
  rowwise() %>% 
  mutate_all(~weight_fn(. ,threshold)) %>%
    pivot_longer(everything(), 
               names_to = "site", 
               names_prefix = "V",
               values_to = "vs") %>% 
  group_by(site) %>% 
  summarise(vs=sum(vs))

node_strength$site <- as.integer(node_strength$site)
node_strength <- node_strength %>% arrange(site)
```

Now we can generate the abundance of breeding waders per survey site

```{r breeding_abun}
# Generate average breeding site count
site_connectivity <- species_sites_sf %>%
  #group_by(Name) %>% 
  #summarise(avg_count = mean(Count)) %>%
  #select(Name, avg_count) %>%
  cbind(.,node_strength = node_strength$vs) %>%
  mutate(area = st_area(.)) %>%
  st_drop_geometry()
```

Fit a gam using a zero-inflated poisson process.

```{r regression}

library(mgcv)
fit1 <- gam(Count ~ s(node_strength, k=5),  
            offset=log(Area_km2), 
            family = ziP(), 
            data = site_connectivity)

plot(fit1, all.terms=F, residuals=T, pch=20) 
summary(fit1)
```

No plot on the response scale

```{r plot_response}
# Plot prediction
library(tidymv)
get_gam_predictions(fit1,node_strength, transform = exp) %>%
  #filter(node_strength > 1, node_strength < 5.5) %>%
  ggplot() +
  geom_line(aes(x=node_strength,y=Count, colour="Count")) +
  geom_ribbon(aes(ymin=CI_lower, ymax=CI_upper, x=node_strength, fill = "se"), alpha = 0.3) +
  scale_colour_manual("",values="blue")+
  scale_fill_manual("",values="grey12")
```