## Information Theory (IT) covariates 

GAMs for breeding wader abundance response across the five different species were generated from 2002–10 and 2011-19. They included 5 IT predictor variables (covariates), and the model parameters for all species across the two response periods can be seen in Table \@ref(tab:tableGamParamsIT). A third model was generated by taking the abundance response from the first two models to generate the ratio of population change between the two response periods.

### Histograms of Information theory covariates

Histograms of IT covariates using the EUNIS landscape categorisation, across all of Shetland are shown in figure \@ref(fig:itHisto). The *marginal entropy* for the Shetland landscape is approximately normally distributed, indicating that habitat within the Shetland landscape is spatially diverse but that very low and highly diverse habitat within Shetland are relatively rare. The mode of the *conditional entropy* is relatively low with a distribution that shows significant positive skew; this suggests that the Shetland landscape has relatively low geometric intricacy. This arises when cells of one category within a landscape raster are predominantly adjacent to cells of the same category. The overall spatio-thematic complexity is measured by *joint entropy*. This can be thought of as quantifying the uncertainty in determining the habitat type of a focus cell and an adjacent cell. For Shetland, joint entropy appears to be approximately normally distributed. This indicates that habitat with very high or low spatio-thematic complexity is relatively rare on Shetland. Due to the spatial autocorrelation, the value of *mutual information* tends to grow with a diversity of the landscape (marginal entropy). To adjust this tendency, it is possible to calculate *relative mutual information* by dividing the mutual information by the marginal entropy. Relative mutual information always has a range between 0 and 1, and quantifies the degree of aggregation of spatial habitat. It can be seen that for Shetland, relative mutual information is distributed with significant negative skew. This implies that habitat types across Shetland are predominantly aggregated - small relatively information values indicate significant fragmentation in landscape habitat patterns.

```{r itHisto, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Histograms of Information Theory covariates across all Shetland OS 1km squares'}
# Plot histogram
histo_covars %>%
  # pivot long by IT metrics
  pivot_longer(cols=c('Marginal entropy':'Relative mutual informaton'), 
               names_to = "covar",
               values_to = "value") %>%
  # Remove zero values as this makes it hard to see the distribution
  filter(value >0) %>%
  ggplot(aes(x = value)) + 
  # plot the histogram
  stat_bin(color = "black", bins=20) +
  theme_minimal() +
  labs(x = "Information Theory covariate", y = "Frequency") +
  facet_wrap(~covar, ncol=2, scales="free")
```

### Histograms of Information Theory covariates for surveyed squares only

1km squares surveyed as part of the Shetland BBS were used to generate IT covariates histograms using the EUNIS landscape categorisation, as shown in Figure \@ref(fig:itSBBSHisto). Here we can see that the *conditional entropy* and the *marginal entropy* across all surveyed squares had a mode that was significantly higher than the Shetland wide values shown in Figure \@ref(fig:itHisto). There is also significantly less negative negative skew in the *relative mutual information* of the surveyed squares. 

```{r itSBBSHisto, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Histograms of Information Theory covariates across SBBS surveyed sqaures only '}
# Load covar data for surveyed BBS squares
bbs_covars <- surveyed_PLAN_NO %>% 
  dplyr::left_join(histo_covars, by="PLAN_NO") %>%
  na.omit() %>%
  st_drop_geometry()

# Plot histogram
bbs_covars %>%
  # pivot long by sIT metrics
  pivot_longer(cols=c('Marginal entropy':'Relative mutual informaton'), 
               names_to = "covar",
               values_to = "value") %>%
  # Remove zero values as this makes it hard to see the distribution
  filter(value >0) %>%
  ggplot(aes(x = value)) + 
  # plot the histogram
  stat_bin(color = "black", bins=20) +
  theme_minimal() +
  labs(x = "Environmental covariate", y = "Frequency") +
  facet_wrap(~covar, ncol=2, scales="free")

```

### Information Theory covariates abundance response model

```{r itGam,  message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Information Metrics as covariates for abundance response'}

it_wader_df <- wader_covars_df %>% 
  # Structure data by covariate and then group
  # pivot long by sIT metrics
  pivot_longer(cols=c('Marginal entropy':'Relative mutual informaton'), 
               names_to = "covariate",
               values_to = "value") %>%
  group_by(period, Species, covariate) %>%
  nest() %>%
  mutate(model = map2(.x = data,
                      .y = period,
                      # Generate a univariate GAM model for wader count against covariate (value).
                      # Repeat for each combination of wader and covariate 
                      ~ gam(count_sm ~ s(X,Y, k=20) + value,
                                      # Apply a different offset depending on whether this gam
                                      # models population change or only covariate response
                                      offset = case_when(.y == "2002-2019" ~ log(count_offset), 
                                                         .y != "2002-2019" ~ log(d) + log(9)),
                                      family = poisson(link="log"),
                                      method = "REML",
                                      data = .x)),
    # Summarise model config for reporting
  model_smry = map(model, ~tidy(.,parametric=TRUE)  )) %>%
  unnest(cols=model_smry) %>%
  filter(term != "(Intercept)") %>%
  dplyr::select(-term)

```

Here we generate GAMs using IT metrics as covariates against breeding wader abundance data. Figures \@ref(fig:itHeatMap200210) and \@ref(fig:itHeatMap201119) summarise the associations between the abundance response and IT covariates used in the univariate GAMs, for the periods 2002-10 and 2011-19 respectively.

```{r itHeatMap200210, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of associations between breeding wader density and IT covariates, between 2002 and 2010 inclusive'}
it_species_assoc_2002_10 <- data.frame(
  Covariate = c('Marginal entropy',
                'Conditional entropy',
                'Joint entropy',
                'Mutual information',
                'Relative mutual inforation'),
  CU = c("NS","NS","NS","NS","Pos"),
  L =  c("Pos","NS","NS","Pos","Pos"),
  OC = c("NS","Neg","NS","Pos","Pos"),
  RK = c("Pos","NS","Pos","Pos","Pos"),
  SN = c("Neg","Neg","Neg","Neg","Pos")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

it_species_assoc_2002_10 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("grey", "red", "blue")) +
  theme_minimal()
```

```{r itHeatMap201119, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of associations between breeding wader density and IT covariates, between 2011 and 2019 inclusive'}
it_species_assoc_2011_19 <- data.frame(
  Covariate = c('Marginal entropy',
                'Conditional entropy',
                'Joint entropy',
                'Mutual information',
                'Relative mutual inforation'),
  CU = c("Pos","Pos","Pos","Pos","NS"),
  L =  c("Pos","Pos","Pos","Pos","NS"),
  OC = c("Pos","NS","Pos","Pos","NS"),
  RK = c("Pos","Pos","Pos","Pos","NS"),
  SN = c("Neg","Neg","Neg","Neg","Pos")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

it_species_assoc_2011_19 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("grey", "red", "blue")) +
  theme_minimal()
```
Table \@ref(tab:tableGamParamsITAbundance) shows the GAM parameters generated by fitting the model to the data.

```{r tableGamParamsITAbundance, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='GAM parameter estimates, standard errors, z and P values from models incorporating parameteric terms associated with abundance (in 2002-2010 and 2011-2019) and Information Theory covariates'}
# create table of results of gam fit for each species - abundance response only
library(kableExtra)
it_wader_df %>%
    dplyr::select(Species, period,covariate,estimate:p.value) %>%
    filter(period != "2002-2019") %>%
    kableExtra::kable(
      align = c('c','l','l','c','c','c','c'),
      col.names = c("Species","Response","Covariate",	"Estimate",	"se", "z", "p-value")) %>%
    kable_styling(full_width = F, font_size = 10) %>%
    column_spec(1, bold = T) %>%
    collapse_rows(columns = 1:2, valign = "top") 

```

Plots for abundance response against IT covariates are shown in Figure \@ref(fig:itResponsePlots).

```{r itResponsePlots, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Plots show abundance response to information theory covariates for a GAM'}
# Generate prediction data so that we can plot parametric terms
it_wader_df <- it_wader_df %>% 
  mutate(model_pred = map2(
    .x = model, 
    .y = covariate, 
    # A function to generate a response with parametric covariate only
    ~tidymv::get_gam_predictions(model = .x,
                                 series = .y, 
                                 # We want the response not link
                                 transform = exp)))

# Generate plot for each model
it_wader_df <- it_wader_df %>%
  # Use a column to store plots
  mutate(
    plots = pmap(
      list(w = Species , c = covariate, cs = p.value, m = model_pred, d = period),
      function(w,c,cs,m,d){
         ggplot(data = m) +
         # Plot parametric coefficient - red for plots with p <= 0.05
         geom_line(aes(x=value,y=count_sm), colour = if_else(cs>0.05, 1,2)) + 
         # Plot confidence interval
         geom_ribbon(aes(x =value, ymin=CI_lower,ymax=CI_upper), alpha=0.2) +
         # x axis label set to covariate in model  
         xlab(c) +
         # y axis label set to wader name  
         ylab(paste(w, if_else(d == "2002-2019", "pop chg ratio", "density"), sep = " ")) +
         theme_bw(base_size = 6) +
         theme(legend.position = 'none')
      }))

# Library to arrange plots into pages
library(gridExtra)

# Generate output
it_wader_df %>% 
  # We dont want the population change plots
  filter(period !="2002-2019") %>%
  ungroup %>% 
  dplyr::select(period, plots) %>%
  # Group plots into date ranges
  group_by(period) %>%
  # Map plotting over group
  group_map(
    ~marrangeGrob(.x$plots, 
                  nrow =5, 
                  ncol = 2,
                  # .y gives the key of the grouping variable
                  top = as.character(.y)
                  ))

```

### Population change model against IT covarirates

By using the response of the 2002-2010 wader densities as the offset for the 2011-19 densities, a third series of GAMs were fitted to show the ratio of population change in response to environmental covariates. This is summarised in Figure \@ref(fig:itHeatMapPopChg). It can be seen that there are no statistically significant results for Redshank or Snipe, or for marginal entropy as a covariate.

```{r itHeatMapPopChg, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap='Summary of associations between breeding wader density and IT covariates, between 2011 and 2019 inclusive'}
it_pop_chg_2011_19 <- data.frame(
  Covariate = c('Marginal entropy',
                'Conditional entropy',
                'Joint entropy',
                'Mutual information',
                'Relative mutual inforation'),
  CU = c("NS","Pos","Pos","NS","NS"),
  L =  c("NS","Pos","Pos","NS","Neg"),
  OC = c("NS","Pos","Pos","NS","Neg"),
  RK = c("NS","NS","NS","NS","NS"),
  SN = c("NS","NS","NS","NS","NS")
  )%>%
  pivot_longer(cols = CU:SN, names_to = "Species", values_to ="Association")

it_pop_chg_2011_19 %>% 
  ggplot(mapping = aes(x = Species, y = Covariate, fill = Association)) +
  geom_tile(colour="white") + 
  scale_fill_manual(values=c("grey", "red", "blue")) +
  theme_minimal()
```
Table \@ref(tab:tableGamParamsITPopChg) shows the GAM parameters generated by fitting the population change model to the data.

```{r tableGamParamsITPopChg, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='GAM parameter estimates, standard errors, z and P values from models incorporating parameteric terms associated with abundance (in 2002-2010 and 2011-2019)'}
# create table of results of gam fit for each species - population change only
it_wader_df %>%
    dplyr::select(Species, period, covariate,estimate:p.value) %>%
    filter(period == "2002-2019") %>%
    kableExtra::kable(
      align = c('c','l','l','c','c','c','c'),
      col.names = c("Species","Period", "Covariate","Estimate",	"se", "z", "p-value")) %>%
    kable_styling(full_width = F, font_size = 10) %>%
    column_spec(1, bold = T) %>%
    collapse_rows(columns = 1:2, valign = "top") 
```

Plots for population change response against IT covariates are shown in Figure \@ref(fig:itPopChgPlots).

```{r itPopChgPlots, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.asp= 1, fig.cap='Plots show population change to information theory covariates for a GAM, across all wader species' }
# Generate output
it_wader_df %>% 
  # We dont want the population change plots
  filter(period =="2002-2019") %>%
  ungroup %>% 
  dplyr::select(period, plots) %>%
  # Group plots into date ranges
  group_by(period) %>%
  # Map plotting over group
  group_map(
    ~marrangeGrob(.x$plots, 
                  nrow =5, 
                  ncol = 2,
                  # .y gives the key of the grouping variable
                  top = as.character(.y)
                  ))

```