## Wader abundance trends

```{r load_libs_abund, include=FALSE}
library(tidymodels)
library(tidyverse)
library(sf)
```

<!-- 
Code to generate Random Forest regression model to estimate 
population abundance for each wader species.

Method is as follows:

1 - Load data and generate suitable format
2 - Split into training and test data sets
3 - Preprocess data
-->

```{r load_species_data, message=FALSE, warning=FALSE, include=FALSE}
# load BBS data for all species and years
curlew        <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Curlew.xlsx", col_names = T)
lapwing       <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Lapwing.xlsx", col_names = T)
oystercatcher <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/oystercatcher.xlsx", col_names = T)
redshank      <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Redshank.xlsx", col_names = T)
snipe         <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Snipe.xlsx", col_names = T)

# join all BBS data into one dataframe
bbs_df <- rbind(curlew,
                lapwing,
                oystercatcher,
                redshank,
                snipe)

# Now pivot into long format, and clean up data that have no breeding observations or did not have a survey.
bbs_df <- bbs_df %>%
  # We aren't interested in non-breeding birds that are present.
  mutate_if(is.character, ~ifelse(.=="P", 0, .)) %>%
  mutate_if(is.character, ~ifelse(.=="p", 0, .)) %>%
  # Mainland observations only
  filter(str_detect(PLAN_NO,"^HU") | str_detect(PLAN_NO, "^HP")) %>%
  # Pivot data into long format
  pivot_longer(cols = c(`2002`:`2019`), names_to = "Year", values_to = "Count") %>%
  # Change type for number columns
  mutate_at(.vars = c("Year", "Count"), ~as.numeric(.)) %>%
  # Any count with NA means no survey
  na.omit()

#load covariate data
covars_sf <- readRDS("~/Documents/GitHub/shetlandwaders/data/covars_df") %>%
  st_centroid() %>% 
  cbind(., sf::st_coordinates(.)) 

# Add centroid for each sqaure
covars_df <- covars_sf %>% 
    st_drop_geometry()

# Create Waders data from for all observation data
wader_covars_df <- left_join(bbs_df, covars_df, by="PLAN_NO") %>%
  na.omit()

# Create adjust count to take account of bbs square area and species detectability
detect_df <- data.frame(
  Species = c("OC", "L", "RK", "CU", "SN"),
  d = c(0.803, # oystercatcher 
        0.723, # lapwing
        0.667, # redshank
        0.831, # curlew
        0.723 # snipe
        )) 

# Join detectability
wader_covars_df <- wader_covars_df %>% left_join(detect_df)
# Adjust count
wader_covars_df <- wader_covars_df %>% 
  mutate(Count = Count*area/(1000*1000*d))

# Get rid of all covars that are not needed for population modelling
wader_covars_df <- wader_covars_df %>% 
  dplyr::select(Species, Year, Count, elev, D, E, `F`, wgt_pH, wgt_orgCarbon,AWC, dist_sea, bare_pc_cvr, imp_gr_pc_df,X,Y)

```

```{r data_sampling, message=FALSE, warning=FALSE, include=FALSE}
# Split data between analysis and assess - 60%/40%
set.seed(123)
species_df <- wader_covars_df %>% 
  nest(-Species) %>% 
  mutate(split       = map(.x = data,  ~initial_split(.x, prop=0.6)),
         sp_training = map(.x = split, ~training(.x)),
         sp_testing  = map(.x = split, ~testing(.x)))

# Create recipe for pre-processing of data. Apply a recipe to the training data whereby:
# * Removes covariates that have large absolute correlations with other covariates
# * Normalize numeric data to have a mean of zero
# * Normalize numeric data to have a standard deviation of one
species_df <- species_df %>%
  mutate(recipe = map(
    .x=sp_training,
    ~recipe(.x, Count ~.) %>%
      # Remove covariates that are 80% correlated
      step_corr(all_predictors(), threshold = 0.8) %>%   
      step_center(all_predictors(), -all_outcomes()) %>%
      step_scale(all_predictors(), -all_outcomes()))) 
```

<!-- Create a model specification for a random forest where we will tune mtry (the number of predictors to sample at each split) and min_n (the number of observations needed to keep splitting nodes). -->

```{r tune_spec, message=FALSE, warning=FALSE, include=FALSE}
# Set-up a specification for tuning - same for each training dataset
tune_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
  ) %>%
  set_mode("regression") %>%
  set_engine("ranger",keep.inbag=T)

# Put it all together in a workflow
species_df <- species_df %>%
  mutate(
    tune_wf = map(
      .x = recipe,
      ~workflow() %>% 
         add_recipe(.x) %>% 
         add_model(tune_spec)))
```

<!-- Now we can tune the hyperparameters for a random forest model. First, let’s create a set of cross-validation resamples to use for tuning. Then run a model for each sample dataset. -->

```{r hyperparam_tuning, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
# Create x10 cross-validation resamples for each species dataset
set.seed(234)
species_df <- species_df %>%
  mutate(
    sp_folds = map(
      .x=sp_training,
      # Use 5-fold valdiation
      ~vfold_cv(.x, v=5)))

# Run the workflow
set.seed(345)
species_df <- species_df %>%
  mutate(
    tune_res = map2(
      .x=tune_wf,
      .y=sp_folds,
      # Generate models for tuning
      ~tune_grid(
        .x,
        resamples=.y,
        # Choose 10 random points in the tune grid
        grid=10)))

```

The tuning grid only samples 10 points across the five different folds. Plot the results.

```{r show_results, echo=FALSE, message=FALSE, warning=FALSE}
# Generate metrics for each species tuning run
# Then map into plots of tuning parameter ranges
species_df <- species_df %>% 
  mutate(tune_plot=map2(
    .x=tune_res,
    .y=Species,
    ~collect_metrics(.x) %>%
      filter(.metric == "rmse") %>%
      dplyr::select(mean, min_n, mtry) %>%
      pivot_longer(min_n:mtry,
        values_to = "value",
        names_to = "parameter") %>%
      # Map to plots
      ggplot(aes(value, mean, color = parameter)) +
             geom_point(show.legend = FALSE) +
             facet_wrap(~parameter, scales = "free_x") +
             # Add Species name to each plot
             labs(title = .y, x = NULL, y = "rmse")))

# Plot the results
species_df$tune_plot      

# Drop the plot data column
species_df <- species_df %>% dplyr::select(-tune_plot)
```
<!-- Now refine the tuning per species -->

```{r reg_grid_tune, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
# Redshank tuning grid
RK_grid <- grid_regular(
  mtry(range = c(7, 9)),
  min_n(range = c(1, 10)),
  levels = 10)

# Curlew tuning grid
CU_grid <- grid_regular(
  mtry(range = c(3, 6)),
  min_n(range = c(5, 12)),
  levels = 10)

# Oystercatcher tuning grid
OC_grid <- grid_regular(
  mtry(range = c(3, 9)),
  min_n(range = c(1, 10)),
  levels = 10)

# Lapwing tuning grid
L_grid <- grid_regular(
  mtry(range = c(3, 6)),
  min_n(range = c(1, 4)),
  levels = 10)

# Snipe tuning grid
SN_grid <- grid_regular(
  mtry(range = c(2, 4)),
  min_n(range = c(5, 10)),
  levels = 10)

species_grids <- list(CU_grid,
                      L_grid,
                      OC_grid,
                      RK_grid,
                      SN_grid)

#Now try a more targeted tuning process
set.seed(456)
species_df <- species_df %>% mutate(
  retune_res = pmap(
    list(
      tw=tune_wf,
      re=sp_folds,
      gr=species_grids), 
    function(tw,re,gr){
      tune_grid(tw,resamples=re,grid=gr)
    }))
```

Now we can plot it: the rmse for various permutations of `min_n` and `mtry`.

```{r plot_retune_results, echo=FALSE, message=FALSE, warning=FALSE}
# Plot results
# Then map into plots of tuning parameter ranges
species_df <-species_df %>% 
  mutate(retune_plot=map2(
    .x=retune_res,
    .y=Species,
    ~collect_metrics(.x) %>%
      filter(.metric == "rmse") %>%
      mutate(min_n = factor(min_n)) %>%
      # Map to plots
      ggplot(aes(x=mtry, y=mean, color = min_n)) +
         geom_line(alpha = 0.5, size = 1.5) +
         geom_point() +
         labs(title=.y, y = "rmse")))

# Plot parameters of returned models
species_df$retune_plot
# Choose the best model
```

From the plot above we can see which parameters give the best fit, when using root mean squared error as the metric.  For each function the minimum rmse was as follows:

```{r bestByRMSE, echo=FALSE, message=FALSE, warning=FALSE}
species_df$retune_res %>% 
  map_dfr(~collect_metrics(.x)%>%
  filter(.metric =="rmse") %>%
  arrange(.$mean) %>%
    head(1)) %>%
  bind_cols(Species = species_df$Species, .) %>%
  dplyr::select(Species, mtry, min_n,mean) %>%
    knitr::kable(
    col.names = c("Species","mtry","min_n","rmse"),
    booktabs = TRUE
  )



# Select the best model for each species
species_df <- species_df %>%
  mutate(best_rmse = map(
    .x=retune_res,
    ~select_best(.x,"rmse")),
    final_rf = map(
      .x=best_rmse,
      ~finalize_model(tune_spec, .x)))
```

Having selected the best model fit the variable importance was assessed for each model:

The `vip` package can be used to explore the relative importance of different covariates in the model fit.

```{r variable_importance, echo=FALSE, message=FALSE, warning=FALSE}
library(vip)

# Map over dataset to generate variable importance per species
set.seed(11122)
species_df <- species_df %>%
  mutate(vip_plots=pmap(
    list(rf=final_rf,
         r=recipe,
         s=Species),
    function(rf,r,s){
    rf %>% 
      set_engine("ranger",importance = "permutation") %>%
      fit(Count~., prep(r) %>% juice()) %>% 
      # Plot using the VIP library
      vip(geom="point") +
        ggtitle(s) +
        theme_bw()
      })) 

species_df$vip_plots
```

### Final fit

Fit the final model fit over the dataset recipe

```{r final_fit, message=FALSE, warning=FALSE, include=FALSE}
species_df <- species_df %>%
  mutate(final_wf=map2(
    .x=final_rf,
    .y=recipe,
    ~workflow() %>% 
      add_recipe(.y) %>% 
      add_model(.x)))

# Fit the final model to the training data
species_df <- species_df %>%
  mutate(fit = map2(
    .x=final_wf,
    .y=sp_training,
    ~fit(.x,.y)))
```

### Generate population estimate over time

Predict over all Shetland BBS squares

```{r simple_workflow_test, echo=FALSE, message=FALSE, warning=FALSE,cache = TRUE}
# Create new dataframe to store results
results_df <- expand.grid(
  Species = c("CU","L","OC","RK","SN"),
  Year = seq(2002,2019,1))

results_df <- species_df %>% 
  dplyr::select(Species,fit) %>% 
  left_join(results_df, by="Species")

# Function to calculate the population for a given year
# and a given set of covariates
calc_annual_pop <- function(m, d, y){
  pop = predict(m, d) %>% sum()
  ci  = predict(m, type="conf_int", d) %>% as_tibble()
  out <- list(Pop=pop,
              l_ci = ci %>% pluck(".pred_lower") %>% sum,
              u_ci = ci %>% pluck(".pred_upper") %>% sum)
  return(out)
}


# Calculate population by year, across all species
results_df <- results_df %>% 
  mutate(pop = map2_df(
    .x=Year,
    .y=fit,
    ~calc_annual_pop(.y, 
                     covars_df %>% 
                     na.omit %>% 
                     bind_cols(Count=NA, Year=.x), .x))) 


# Plot species population curves
results_df %>%
  ggplot(aes(y=pop$Pop,x=Year)) +
  geom_line(color='#00aaff') +
  geom_ribbon(aes(ymin = pop$l_ci, ymax=pop$u_ci), fill='gray90') +
  geom_line(color='#00aaff') +
  ylim(0,10000) +
  labs(title = "Breeding wader abundance ", 
       y = "Abundance",
       x = "Year") +
  facet_wrap(~Species, ncol = 3, scales = "free_y") +
  theme_minimal()

``` 

## Spatial abundance distriution

Now we can generate spatial abundance distribution plots.

```{r spatial_distributions, echo=FALSE, message=FALSE, warning=FALSE}
# Generate species abundance distribution data structure
sad <- results_df %>% dplyr::select(Species,fit,Year)
# Covariates over Shetland landmass only
bbs_covars <- covars_sf %>% 
  na.omit 

# Now for each model make a density prediction for each BBS square
plots_df <- pmap_df(
  list(s=sad$Species, y=sad$Year, m=sad$fit),
  function(s,y,m){
    p <- predict(
      m, 
      cbind(bbs_covars, Year=y))
    # Create structure to return dta
    l <- tibble(species=s, year =y, p, X = bbs_covars$X, Y = bbs_covars$Y)
    return(l)
    }
  ) 

# Plot the species abundance distribution.
# Here we plot 2002 v 2019 only
library(viridis)
plots_df %>% filter(year == 2002) %>%
      ggplot() +
      geom_tile(aes(x=X, y=Y, fill=.pred, width = 1000, height = 1000)) +
      coord_equal() +
      labs(fill="Density") +
      scale_fill_distiller(palette = "Spectral") +
      #scale_fill_viridis_c(trans = "sqrt") +
      theme_minimal() +
      labs(title = "Spatial distribution of breeding wader density (count/km2)") +
      facet_grid(~species) +
      theme(axis.text = element_blank(),
            axis.title = element_blank(),
            panel.grid = element_blank())

```
## Net abundance change by species between 2002 and 2019

```{r net_chg_plot, echo=FALSE, message=FALSE, warning=FALSE}
plots_df %>%
  filter(year %in% c(2002,2019)) %>%
  pivot_wider(names_from = c(year), values_from = c(.pred)) %>%
  mutate(net_chg = `2019`-`2002`) %>%
      ggplot() +
      geom_tile(aes(x=X, y=Y, fill=net_chg, width = 1000, height = 1000)) +
      coord_equal() +
      labs(fill="Density Chg") +
      scale_fill_gradient2() +
      #scale_fill_distiller(palette = "Spectral") +
      labs(title = "Change in breeding wader density (count/km2)") +
      theme_minimal() +
      facet_grid(~species) +
      theme(axis.text = element_blank(),
            axis.title = element_blank(),
            panel.grid = element_blank()) 

```
