## Wader abundance trends

A random forest regression model was used to fit an abundance response using all 10 environmental covariates. The parameterisation and results are outlined below.

```{r load_libs_abund, include=FALSE}
library(tidymodels)
library(tidyverse)
library(sf)
```

<!-- 
Code to generate Random Forest regression model to estimate 
population abundance for each wader species.

Method is as follows:

1 - Load data and generate suitable format
2 - Split into training and test data sets
3 - Preprocess data
-->

```{r load_species_data, message=FALSE, warning=FALSE, include=FALSE}
# load BBS data for all species and years
curlew        <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Curlew.xlsx", col_names = T)
lapwing       <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Lapwing.xlsx", col_names = T)
oystercatcher <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/oystercatcher.xlsx", col_names = T)
redshank      <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Redshank.xlsx", col_names = T)
snipe         <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/bbs_species_data/Snipe.xlsx", col_names = T)

# join all BBS data into one dataframe
bbs_df <- rbind(curlew,
                lapwing,
                oystercatcher,
                redshank,
                snipe)

# Now pivot into long format, and clean up data that have no breeding observations or did not have a survey.
bbs_df <- bbs_df %>%
  # We aren't interested in non-breeding birds that are present.
  mutate_if(is.character, ~ifelse(.=="P", 0, .)) %>%
  mutate_if(is.character, ~ifelse(.=="p", 0, .)) %>%
  # Mainland observations only
  filter(str_detect(PLAN_NO,"^HU") | str_detect(PLAN_NO, "^HP")) %>%
  # Pivot data into long format
  pivot_longer(cols = c(`2002`:`2019`), names_to = "Year", values_to = "Count") %>%
  # Change type for number columns
  mutate_at(.vars = c("Year", "Count"), ~as.numeric(.)) %>%
  # Any count with NA means no survey
  na.omit()

#load covariate data
covars_sf <- readRDS("~/Documents/GitHub/shetlandwaders/data/covars_df") %>%
  st_centroid() %>% 
  cbind(., sf::st_coordinates(.)) 

# Add centroid for each sqaure
covars_df <- covars_sf %>% 
    st_drop_geometry()

# Create Waders data from for all observation data
wader_covars_df <- left_join(bbs_df, covars_df, by="PLAN_NO") %>%
  na.omit()

# Create adjust count to take account of bbs square area and species detectability
detect_df <- data.frame(
  Species = c("OC", "L", "RK", "CU", "SN"),
  d = c(0.803, # oystercatcher 
        0.723, # lapwing
        0.667, # redshank
        0.831, # curlew
        0.723 # snipe
        )) 

# Join detectability
wader_covars_df <- wader_covars_df %>% left_join(detect_df)
# Adjust count
wader_covars_df <- wader_covars_df %>% 
  mutate(Count = Count*area/(1000*1000*d))

# Get rid of all covars that are not needed for population modelling
wader_covars_df <- wader_covars_df %>% 
  dplyr::select(Species, Year, Count, elev, D, E, `F`, wgt_pH, wgt_orgCarbon,AWC, dist_sea, bare_pc_cvr, imp_gr_pc_df,X,Y)

```

```{r data_sampling, message=FALSE, warning=FALSE, include=FALSE}
# Split data between analysis and assess - 60%/40%
set.seed(123)
species_df <- wader_covars_df %>% 
  nest(-Species) %>% 
  mutate(split       = map(.x = data,  ~initial_split(.x, prop=0.6)),
         sp_training = map(.x = split, ~training(.x)),
         sp_testing  = map(.x = split, ~testing(.x)))

# Create recipe for pre-processing of data. Apply a recipe to the training data whereby:
# * Removes covariates that have large absolute correlations with other covariates
# * Normalize numeric data to have a mean of zero
# * Normalize numeric data to have a standard deviation of one
species_df <- species_df %>%
  mutate(recipe = map(
    .x=sp_training,
    ~recipe(.x, Count ~.) %>%
      # Remove covariates that are 80% correlated
      step_corr(all_predictors(), threshold = 0.8) %>%   
      step_center(all_predictors(), -all_outcomes()) %>%
      step_scale(all_predictors(), -all_outcomes()))) 
```

<!-- Create a model specification for a random forest where we will tune mtry (the number of predictors to sample at each split) and min_n (the number of observations needed to keep splitting nodes). -->

```{r tune_spec, message=FALSE, warning=FALSE, include=FALSE}
# Set-up a specification for tuning - same for each training dataset
tune_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
  ) %>%
  set_mode("regression") %>%
  set_engine("ranger",keep.inbag=T)

# Put it all together in a workflow
species_df <- species_df %>%
  mutate(
    tune_wf = map(
      .x = recipe,
      ~workflow() %>% 
         add_recipe(.x) %>% 
         add_model(tune_spec)))
```

<!-- Now we can tune the hyperparameters for a random forest model. First, let’s create a set of cross-validation resamples to use for tuning. Then run a model for each sample dataset. -->

```{r hyperparam_tuning, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Create x10 cross-validation resamples for each species dataset
set.seed(234)
species_df <- species_df %>%
  mutate(
    sp_folds = map(
      .x=sp_training,
      # Use 5-fold valdiation
      ~vfold_cv(.x, v=5)))

# Run the workflow
set.seed(345)
species_df <- species_df %>%
  mutate(
    tune_res = map2(
      .x=tune_wf,
      .y=sp_folds,
      # Generate models for tuning
      ~tune_grid(
        .x,
        resamples=.y,
        # Choose 10 random points in the tune grid
        grid=10)))

```

### Tuning the model hyper parameters

A search grid over 10 different folds gave initial results for various permutations of hyper parameters as shown in Appendix \@ref(hyp-params). Each figure shows the results for a particular species. Each hyper-parameter permutation is plotted against the resulting root mean squared error (rmse). 

### Further model hyper parameter tuning

Given the initial results in \@ref(hyp-params) the initial range over which the hyper-parameter search was conducted, was refined further by searching over a revised range according by selecting upper and lower limits for hyper-parameters that gave the lowest rmse as indicated by the initial results. The range used was that which gave the lowest rmse as given in Appendix \@ref(hyp-params). The results of the revised search grid range can be seen in Appendix \@ref(final-fit). Again each plot is for a separate species.

<!-- Now refine the tuning per species -->

```{r reg_grid_tune, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Redshank tuning grid
RK_grid <- grid_regular(
  mtry(range = c(7, 9)),
  min_n(range = c(1, 10)),
  levels = 5)

# Curlew tuning grid
CU_grid <- grid_regular(
  mtry(range = c(3, 6)),
  min_n(range = c(5, 12)),
  levels = 5)

# Oystercatcher tuning grid
OC_grid <- grid_regular(
  mtry(range = c(3, 9)),
  min_n(range = c(1, 10)),
  levels = 5)

# Lapwing tuning grid
L_grid <- grid_regular(
  mtry(range = c(3, 6)),
  min_n(range = c(1, 4)),
  levels = 5)

# Snipe tuning grid
SN_grid <- grid_regular(
  mtry(range = c(2, 4)),
  min_n(range = c(5, 10)),
  levels = 5)

species_grids <- list(CU_grid,
                      L_grid,
                      OC_grid,
                      RK_grid,
                      SN_grid)

#Now try a more targeted tuning process
set.seed(456)
species_df <- species_df %>% mutate(
  retune_res = pmap(
    list(
      tw=tune_wf,
      re=sp_folds,
      gr=species_grids), 
    function(tw,re,gr){
      tune_grid(tw,resamples=re,grid=gr)
    }))
```


From the plots in Appensix \@ref(final-fit) we can see which hyper parameters give the best fit, when using root mean squared error as an evaluation metric. It can be seen that the model fit for Snipe has the largest RMSE. whilst Redshank has the lowest. For each species the minimum rmse result for the best model fit, together with the associated hyper parameters (`trees`= 1000 for all models) is shown in Table \@ref(tab:bestByRMSE).

```{r bestByRMSE, echo=FALSE, message=FALSE, warning=FALSE}

# Load cached data
species_df <- readRDS("~/Documents/GitHub/shetlandwaders/data/cacheddata/abun_ml_2")

species_df$retune_res %>% 
  map_dfr(~collect_metrics(.x)%>%
  filter(.metric =="rmse") %>%
  arrange(.$mean) %>%
    head(1)) %>%
  bind_cols(Species = species_df$Species, .) %>%
  dplyr::select(Species, mtry, min_n, mean) %>%
    knitr::kable(
    col.names = c("Species","mtry","min_n","rmse"),
    booktabs = TRUE,
    caption = 'Lowest RMSE for best fitting model, for each species of breeding wader'
  )
```

```{r selectBestRMSE, message=TRUE, warning=TRUE, include=FALSE}

# Select the best model for each species
species_df <- species_df %>%
  mutate(best_rmse = map(
    .x=retune_res,
    ~select_best(.x,"rmse")),
    final_rf = map(
      .x=best_rmse,
      ~finalize_model(tune_spec, .x)))
```

### Variable importance in model fit

Having selected the best model fit the variable importance for each species was assessed. The `r` package `vip` [@vip] was used to explore the relative importance of different covariates in the model fit. The results are shown in Appendix \@ref(vip-plots).

```{r variableImportance, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Load variable importance
library(vip)

# Map over dataset to generate variable importance per species
set.seed(11122)
species_df <- species_df %>%
  mutate(vip_plots=pmap(
    list(rf=final_rf,
         r=recipe,
         s=Species),
    function(rf,r,s){
    rf %>% 
      set_engine("ranger",importance = "permutation") %>%
      fit(Count~., prep(r) %>% juice()) %>% 
      # Plot using the VIP library
      vip(geom="point") +
        ggtitle(s) +
        theme_bw()
      })) 

species_df$vip_plots
```

It can be seen that pH, X (longitude) and grassland percentage coverage for a given OS 1km square appear to be the most important covariates for predicting abundance in Curlew. For Lapwing, pH, heathland percentage coverage and topsoil organic carbon content are the most important variables. Whilst for Oystercatcher, grassland and heathland percentage cover are almost equivalent in their importance followed by longitude. For Redshank and Snipe, available water capacity and heathland are the most important covariates in predicting abundance.

```{r final_fit, message=FALSE, warning=FALSE, include=FALSE}
# Generate final workflow
species_df <- species_df %>%
  mutate(final_wf=map2(
    .x=final_rf,
    .y=recipe,
    ~workflow() %>% 
      add_recipe(.y) %>% 
      add_model(.x)))

# Fit the final model to the training data
species_df <- species_df %>%
  mutate(fit = map2(
    .x=final_wf,
    .y=sp_training,
    ~fit(.x,.y)))
```

### Breeding wader population estimates from 2002 to 2019 {#pop-estimates}

The random forest regression model was used to predict species abundance over *all* (n=3992) Shetland BBS 1km squares. The model gave a mean estimate together with lower and upper confidence intervals (5% and 95% percentiles respectively), across every year the survey was run (2002 to 2019). The population estimate results are shown in Appendix \@ref(pop-results) and plotted in Figure \@ref(fig:plotAbunResults)

```{r calcAbundanceResults, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Create new dataframe to store results
results_df <- expand.grid(
  Species = c("CU","L","OC","RK","SN"),
  Year = seq(2002,2019,1))

results_df <- species_df %>% 
  dplyr::select(Species,fit) %>% 
  left_join(results_df, by="Species")

# Function to calculate the population for a given year
# and a given set of covariates
calc_annual_pop <- function(m, d, y){
  pop = predict(m, d) %>% sum()
  ci  = predict(m, type="conf_int", d) %>% as_tibble()
  out <- list(Pop=pop,
              l_ci = ci %>% pluck(".pred_lower") %>% sum,
              u_ci = ci %>% pluck(".pred_upper") %>% sum)
  return(out)
}


# Calculate population by year, across all species
results_df <- results_df %>% 
  mutate(pop = map2_df(
    .x=Year,
    .y=fit,
    ~calc_annual_pop(.y, 
                     covars_df %>% 
                     na.omit %>% 
                     bind_cols(Count=NA, Year=.x), .x))) 
```


```{r plotAbunResults,echo=FALSE, message=FALSE, warning=FALSE, fig.align='centre', fig.width= 9, fig.cap='Breeding wader population estimates - number of pairs of breeding waders by species, 2002 to 2019. Grey shaded area indicates upper and lower confidence intervals'}
# Load cached abundance data
results_df <- readRDS("~/Documents/GitHub/shetlandwaders/data/cacheddata/abun_results")
# Plot species population curves
results_df %>%
  ggplot(aes(y=pop$Pop,x=Year)) +
  geom_line(color='#00aaff') +
  geom_ribbon(aes(ymin = pop$l_ci, ymax=pop$u_ci), fill='gray90') +
  geom_line(color='#00aaff') +
  ylim(0,10000) +
  labs(y = "Abundance", x = "Year") +
  facet_wrap(~Species, ncol = 3, scales = "free_y") +
  theme_minimal()
```

Across the years 2002 to 2019 the abundance of breeding wader pairs across all species appear to have decreased, with the exception of Snipe. The most significant decline was for Lapwing. Note that the confidence intervals for Snipe are highly variable in certain years. Table \@ref(tab:popChgTable) shows the change in breeding wader abundance by species between 2002 and 2019.

```{r popChgTable, echo=FALSE,  message=FALSE, warning=FALSE}
popChgTable <- data.frame(
  species = c("CU","L","OC","RK","SN"),
  y2002   = c(4597, 3474, 5269, 2390, 6043),
  y2019   = c(4088, 2638, 4760, 2248, 7391))

popChgTable$chg <- ((popChgTable$y2019 - popChgTable$y2002)/popChgTable$y2002 * 100) %>% round(1)

popChgTable %>% 
    knitr::kable(
    col.names = c("Species","2002","2019","% Change"),
    booktabs = TRUE,
    caption = 'Change in breeding wader abundance across all species, between the years 2002 and 2019')
```

### Lapwing abundance association with grassland holdings {#lapwing-grassland}

Data from 2002 to 2017 for grassland holdings in hectares categorised as: tillage, grassland less than five years old (reseeded grassland), grassland five years or greater and private grazing were tested for normality, and found to be log-normally distributed. A linear regression was plotted for Lapwing abundance versus each grassland categorisation (in Hectares) as shown in Figure \@ref(fig:lapwingGrasslandPlot).

```{r loadAgdata, message=FALSE, warning=FALSE, include=FALSE}
# Load data
agri_df <- readxl::read_xlsx("~/Documents/GitHub/shetlandwaders/data/agri_census.xlsx", col_names = T)
wader_pop_df <- results_df %>% dplyr::select(-fit)

# Fix missing data - straight line impute
agri_df$tillage[[10]] <- (agri_df$tillage[[9]] - agri_df$tillage[[11]])/2 + agri_df$tillage[[11]]
agri_df$grass_lt_5_y[[10]] <- (agri_df$grass_lt_5_y[[9]] - agri_df$grass_lt_5_y[[11]])/2 + agri_df$grass_lt_5_y[[11]]
agri_df$grass_gt_5_y[[10]] <- (agri_df$grass_gt_5_y[[9]] - agri_df$grass_gt_5_y[[11]])/2 + agri_df$grass_gt_5_y[[11]]
agri_df$private_grazing[[10]] <- (agri_df$private_grazing[[9]] - agri_df$private_grazing[[11]])/2 + agri_df$private_grazing[[11]]
agri_df <- agri_df %>% dplyr::select(-common_grazing)

# join agri and population data
joined_df <- results_df %>%
  dplyr::select(Year, Species) %>% 
  bind_cols(wader_pop_df$pop) %>%
  dplyr::select(Species, Year, Pop) %>%
  left_join(.,agri_df, by=c("Year"="year"))
```


```{r lapwingGrasslandPlot, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.asp = 1, fig.cap='Association between the size of grassland categories on Shetland and Lapwing population abundancebetween 2002 and 2017. Agricultural statistics taken from Scottish Agricultural Survey'}

# Select Lapwing only
lapwing_df <- joined_df %>% 
  na.omit %>%
  pivot_longer(cols=tillage:private_grazing,
               names_to = "crop", 
               values_to = "hectares") %>% 
  filter(Species == "L") %>%
  mutate(crop = case_when(
    crop == "tillage" ~ "Tillage",
    crop == "grass_lt_5_y" ~ "Grass < 5y",
    crop == "grass_gt_5_y" ~ "Grsss >= 5y",
    crop == "private_grazing" ~ "Private grazing"
  ))

# Plot linear regression association between Lapwing abundance and hectares for each
# grassland category.
lapwing_df %>%
  ggplot(aes(x=hectares, y=Pop)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = y~x) +
  facet_wrap(~crop, scales = "free", ncol=2) +
  theme_minimal()
```
Figure \@ref(fig:lapwingGrasslandPlot) shows that Lapwing population size is positively associated with the size of grassland that is less than five years old (that grassland which has been reseeded) and tillage (land prepared for spring crops). Figure \@ref(fig:lapwingGrasslandPlot) also shows that Lapwing population abundance is negatively associated with grassland that is not reseeded (Grass >= 5y).

The `r` package `gamlss` [@gamlss] was used to fit a log-normal model for Lapwing population abundance with grassland less than 5 years old as a covariate. The grassland < 5 years old was a statistically significant covariate (p<0.0001), and the residuals of the model were approximately normally distributed as shown in Figure \@ref(fig:fitLapwingModel).

```{r fitLapwingModel, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap='Residuals for model fit of Lapwing population estimates to Shetland improved grassland less than five years old. Agricultural statistics taken from Scottish Agricultural Survey'}
# Load library
library(gamlss)
# Fit model
m1 <- gamlss(Pop~hectares, 
             family = LOGNO(),
             # Look at L population for grassland <5 years
             data= lapwing_df %>% filter(crop == "Grass < 5y"))
# Plot the model information
plot(m1, summaries = F)  
```

## Spatial abundance distriution of breeding waders

The population estimates results in \@ref(pop-estimates) are based on a spatial prediction. As such they can be plotted to show how species abundance is spatially distributed across Shetland. Figure \@ref(fig:spatialDistributions) shows the abundance distribution, per 1km$^2$ (density) for each species in 2019. 

```{r spatialDistributions, echo=FALSE, message=FALSE, warning=FALSE,fig.width=10, fig.height=6,fig.align='center', fig.cap='Spatial abundance distribution of breeding waders for 2019'}
# Generate species abundance distribution data structure
sad <- results_df %>% dplyr::select(Species,fit,Year)
# Covariates over Shetland landmass only
bbs_covars <- covars_sf %>% 
  na.omit 

# Now for each model make a density prediction for each BBS square
plots_df <- pmap_df(
  list(s=sad$Species, y=sad$Year, m=sad$fit),
  function(s,y,m){
    p <- predict(
      m, 
      cbind(bbs_covars, Year=y))
    # Create structure to return data
    l <- tibble(species=s, year =y, p, X = bbs_covars$X, Y = bbs_covars$Y)
    return(l)
    }
  ) 

# Plot the species abundance distribution.
# Here we plot 2019 only
library(viridis)
plots_df %>% filter(year == 2019) %>%
      ggplot() +
      geom_tile(aes(x=X, y=Y, fill=.pred, width = 1000, height = 1000)) +
      coord_equal() +
      labs(fill="Density") +
      #scale_fill_distiller(palette = "Spectral") +
      scale_fill_viridis_c(trans = "sqrt") +
      theme_minimal() +
      facet_grid(~species) +
      theme(axis.text = element_blank(),
            axis.title = element_blank(),
            panel.grid = element_blank(),
            legend.position="bottom",
            panel.background = element_rect(fill = 'light grey', colour = 'dark grey'))

```

It can be seen that Snipe are widespread, and that Curlew and Oystercatcher are significantly present on the western side of Shetland, as suggested by the variable importance plots in Appendix \@ref(vip-plots). Lapwing and Redshank have the lowest population densities and seem to be concentrated in the south west of Shetland.

### Net spaital abundance change by species, between 2002 and 2019 

Given the spatial abundance distribution for 2002 and 2019, it is possible to plot the net change in breeding wader abundance, for each 1km$^2$, between the two years. This is shown in Figure \@ref(fig:netChgPlot).

```{r netChgPlot, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=10, fig.height=5,fig.cap='Change in breeding wader density (count/km2) between 2002 and 2019 on Shetland',fig.height=6}
plots_df %>%
  filter(year %in% c(2002,2019)) %>%
  pivot_wider(names_from = c(year), values_from = c(.pred)) %>%
  mutate(net_chg = `2019`-`2002`) %>%
      ggplot() +
      geom_tile(aes(x=X, y=Y, fill=net_chg, width = 1000, height = 1000)) +
      coord_equal() +
      labs(fill="Density Chg") +
      scale_fill_gradient2() +
      #scale_fill_distiller(palette = "Spectral") +
      theme_minimal() +
      facet_grid(~species) +
      theme(axis.text = element_blank(),
            axis.title = element_blank(),
            panel.grid = element_blank(),
            legend.position="bottom",
            panel.background = element_rect(fill = 'light grey', colour = 'dark grey')) 

```

It can be seen that the drop in abundance as shown in Appendix \@ref(pop-results) is reflected in the net change plots in Figure \@ref(fig:netChgPlot). 

The initial analysis of the status of surveyed 1km squares (see Figure \@ref(fig:aggPopChg)) suggested that Lapwing had undergone a decline, and Figure \@ref(fig:netChgPlot) shows that this has occurred in the south west and north east over the two analysis periods. Oystercatcher appear to have undergone a significant decline in the north east of Shetland (the islands of Unst and Fetlar). Where as Snipe appear to have increased in most upland areas of  Shetland, and in particular on the Fetlar RSPB reserve.
